### 2019-06-06 README.Rmd
Use README.Rmd to provie examples in the landing page

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# experDesign

<!-- badges: start -->
[![Travis build status](https://travis-ci.org/llrs/experDesign.svg?branch=master)](https://travis-ci.org/llrs/experDesign)
[![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/llrs/experDesign?branch=master&svg=true)](https://ci.appveyor.com/project/llrs/experDesign)
[![Coverage status](https://codecov.io/gh/llrs/experDesign/branch/master/graph/badge.svg)](https://codecov.io/github/llrs/experDesign?branch=master)
[![lifecycle](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![Project Status: Active - The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active)
<!-- badges: end -->

The goal of experDesign is to help you decide which samples go in which batch, 
reducing the potential batch bias when analyzing.

## Installation

You can install the development version from [GitHub](https://github.com/) with:

```r
# install.packages("devtools")
devtools::install_github("llrs/experDesign")
```
## Example

This is a basic example which shows you how to solve a common problem:

```{r design}
library("experDesign")
metadata <- expand.grid(height = seq(60, 80, 5), 
                        weight = seq(100, 300, 50),
                        sex = c("Male","Female"))
head(metadata, 15)
d <- design(metadata, 24)
# It is a list but we can convert it to a vector with:
batch_names(d)

```

Naively one would either fill some batches fully or distribute them not evenly 
(the  first 17 packages toghether, the next 17 and so on). This solution ensures 
that the data is randomized.

If you need space for replicates to control for batch effect you can use:

```{r replicates}
r <- replicates(metadata, 24, 5)
r
```

Which seeks as controls the most diverse values and adds them to the samples 
distribution.

# Previous work

The [OSAT](https://bioconductor.org/packages/OSAT/) package handles categorical 
variables but not numeric data. It doesn't work with our data.

The [minDiff](https://github.com/m-Py/minDiff) package reported in [Stats.SE](https://stats.stackexchange.com/a/326015/105234), handles both 
numeric and categorical data. But it can only optimize for two nominal criteria.
It doesn't work for our data.

The [DeclareDesign](https://github.com/DeclareDesign/DeclareDesign) or [OSAT](http://bioconductor.org/packages/OSAT) packages are also relevant

Question in [Bioinformatics.SE](https://bioinformatics.stackexchange.com/q/4765/48)

# Other

Please note that this project is released with a [Contributor Code of Conduct](CODE_OF_CONDUCT.md).
By participating in this project you agree to abide by its terms.

### 2019-06-07 R/QC.R
Make  R CMD check pass by improving documentation

@@ -5,7 +5,7 @@
 #' @inheritParams qcSubset
 #' @inheritParams design
 #' @return A vector with the number of the rows that are selected.
-#' @seealso \code{\link{optimum_size}}
+#' @seealso \code{\link{optimum}}
 #' @export
 #' @examples
 #' metadata <- expand.grid(height = seq(60, 80, 5), weight = seq(100, 300, 50),

### 2019-11-06 R/designer.R
Use auxiliar helper

@@ -48,15 +48,8 @@ design <- function(pheno, size_subset, omit = NULL, iterations = 500) {
     differences <- abs(sweep(subsets, c(1, 2), original_pheno))
     # Add the independence of the categorical values
     subset_ind <- evaluate_independence(i, pheno_o)
-
     # Calculate the score for each subset by variable
-    meanDiff <- apply(differences, 3, function(x) {
-
-      x <- rbind(x, "ind" = 0)
-      x <- insert(x, subset_ind, name = "ind")
-      colSums(x, na.rm = TRUE)/eval_n
-    })
-
+    meanDiff <- mean_difference(differences, subset_ind, eval_n)
     # Minimize the value
     optimize <- sum(rowMeans(abs(meanDiff)))
 

### 2019-11-06 R/utils.R

@@ -33,3 +33,14 @@ check_sizes <- function(size_data, n_batch, size_batch){
   }
   FALSE
 }
+
+
+
+mean_difference <- function(differences, subset_ind, eval_n) {
+  # Calculate the score for each subset by variable
+  apply(differences, 3, function(x) {
+    x <- rbind(x, "ind" = 0)
+    x <- insert(x, subset_ind, name = "ind")
+    colSums(x, na.rm = TRUE)/eval_n
+  })
+}

### 2019-11-14 README.Rmd
Add Github actions to test

@@
 [![Project Status: Active - The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active)
+[![R build status](https://github.com/llrs/experDesign/workflows/R-CMD-check/badge.svg)](https://github.com/llrs/experDesign/actions?workflow=R-CMD-check)
 <!-- badges: end -->
@@
 
-This is a basic example which shows you how to solve a common problem:
-
-```{r design}
+Imagine you have some samples already collected and you want to distributed them in batches:
+```{r experDesign_setup}
 library("experDesign")
@@
 head(metadata, 15)
+```
+
+If you block incorrectly and end up with a group in a single batch we will end up with batch effect.
+In order to avoid this `design` helps you assign each sample to a batch (in this case each batch has 24 samples at most.):
+
+```{r design}
 d <- design(metadata, 24)
@@
 batch_names(d)
-
 ```
@@
 Naively one would either fill some batches fully or distribute them not evenly 
-(the  first 17 packages toghether, the next 17 and so on). This solution ensures 
+(the  first 17 packages together, the next 17 and so on). This solution ensures 
 that the data is randomized.

### 2020-04-20 R/designer.R
Improve documentation

@@ -65,12 +65,13 @@ design <- function(pheno, size_subset, omit = NULL, iterations = 500) {
 #' Design a batch experiment with experimental controls
 #'
 #' To ensure that the batches are comparable some samples are processed in each
-#' batch. This function allows to take into account that effect
+#' batch. This function allows to take into account that effect.
+#' It uses the most different samples as controls as defined with `extreme_cases`.
 #' @inheritParams design
 #' @param controls The numeric value of the amount of technical controls per
 #' batch.
 #' @return A index with some samples duplicated in the batches
-#' @seealso \code{\link{design}}
+#' @seealso \code{\link{design}}, \code{\link{extreme_cases}}
 #' @export
 #' @examples
 #' samples <- data.frame(L = letters[1:25], Age = rnorm(25))

### 2020-04-20 README.Rmd
Add a section on the README clarify the Previous Work section

@@
 If you block incorrectly and end up with a group in a single batch we will end up with batch effect.
-In order to avoid this `design` helps you assign each sample to a batch (in this case each batch has 24 samples at most.):
+In order to avoid this `design` helps you assign each sample to a batch (in this case each batch has 24 samples at most). First we can explore the number of samples and the number of batches:
+
+```{r size}
+size_data <- nrow(metadata)
+size_batch <- 24
+(batches <- optimum_batches(size_data, size_batch))
+# So now the best number of samples for each batch is less than the available
+(size <- optimum_subset(size_data, batches))
+# The distribution of samples per batch
+sizes_batches(size_data, size, batches)
+```
+
+Note that instead of using a whole batch and then leave a single sample on the third distributes all the samples in the three batches that will be needed.
+We cann directly look for the distribution of the samples given our max number of samples per batch:
 
 ```{r design}
-d <- design(metadata, 24)
+d <- design(metadata, size_batch)
 # It is a list but we can convert it to a vector with:
@@
 (the  first 17 packages together, the next 17 and so on). This solution ensures 
-that the data is randomized.
+that the data is randomized. For more random distribution you can increase the number of iterations performed to calculate this distribution.
 
@@
 ```{r replicates}
-r <- replicates(metadata, 24, 5)
+r <- replicates(metadata, size_batch, 5)
+lengths(r)
 r
@@
 Which seeks as controls the most diverse values and adds them to the samples 
-distribution.
+distribution. Note that if the sample is already present on that batch is not added again, that's why the number of samples per batch is different from the design without replicates.
 
@@
 
-The [OSAT](https://bioconductor.org/packages/OSAT/) package handles categorical 
+The CRAN task View of [Experimental Design](https://cran.r-project.org/web/views/ExperimentalDesign.html) includes many packages rellevant for designing an experiment before collecting data, but none of them provides how to manage them once the samples are already collected.
+
+Two packages allow to distribute the samples on batches:
+
+- The [OSAT](https://bioconductor.org/packages/OSAT/) package handles categorical 
 variables but not numeric data. It doesn't work with our data.
 
-The [minDiff](https://github.com/m-Py/minDiff) package reported in [Stats.SE](https://stats.stackexchange.com/a/326015/105234), handles both 
+ - The [minDiff](https://github.com/m-Py/minDiff) package reported in [Stats.SE](https://stats.stackexchange.com/a/326015/105234), handles both 
 numeric and categorical data. But it can only optimize for two nominal criteria.
@@
 
-The [DeclareDesign](https://github.com/DeclareDesign/DeclareDesign) or [OSAT](http://bioconductor.org/packages/OSAT) packages are also relevant
+If you are still designing the experiment and do not have collected any data [DeclareDesign](https://cran.r-project.org/package=DeclareDesign) might be relevant for you.
 
-Question in [Bioinformatics.SE](https://bioinformatics.stackexchange.com/q/4765/48)
+Question in [Bioinformatics.SE](https://bioinformatics.stackexchange.com/q/4765/48) I made before developing the package.
 

### 2020-07-24 README.Rmd
Get ready for submission

@@
 
-The CRAN task View of [Experimental Design](https://cran.r-project.org/web/views/ExperimentalDesign.html) includes many packages rellevant for designing an experiment before collecting data, but none of them provides how to manage them once the samples are already collected.
+The CRAN task View of [Experimental Design](https://CRAN.R-project.org/view=ExperimentalDesign) includes many packages rellevant for designing an experiment before collecting data, but none of them provides how to manage them once the samples are already collected.
 

### 2020-07-24 README.Rmd
Update dictionary

@@
 Note that instead of using a whole batch and then leave a single sample on the third distributes all the samples in the three batches that will be needed.
-We cann directly look for the distribution of the samples given our max number of samples per batch:
+We can directly look for the distribution of the samples given our max number of samples per batch:
 
@@
 
-The CRAN task View of [Experimental Design](https://CRAN.R-project.org/view=ExperimentalDesign) includes many packages rellevant for designing an experiment before collecting data, but none of them provides how to manage them once the samples are already collected.
+The CRAN task View of [Experimental Design](https://CRAN.R-project.org/view=ExperimentalDesign) includes many packages relevant for designing an experiment before collecting data, but none of them provides how to manage them once the samples are already collected.
 

### 2020-07-24 R/entropy.R

@@ -3,7 +3,7 @@
 #' Calculates the entropy of a category. It uses the amount of categories to
 #' scale between 0 and 1.
 #' @param x A character or vector with two or more categories
-#' @return The numeric value of the shannon entropy scaled between 0 and 1.
+#' @return The numeric value of the Shannon entropy scaled between 0 and 1.
 #' @note It omits the \code{NA} if present.
 #' @export
 #' @examples

### 2020-09-01 README.Rmd
Fix link to Code of conduct
Remove the unsafe link

@@
 [![lifecycle](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
-[![Project Status: Active - The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active)
+[![Project Status: Active - The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
 [![R build status](https://github.com/llrs/experDesign/workflows/R-CMD-check/badge.svg)](https://github.com/llrs/experDesign/actions?workflow=R-CMD-check)
@@
 
-Please note that this project is released with a [Contributor Code of Conduct](CODE_OF_CONDUCT.md).
+Please note that this project is released with a [Contributor Code of Conduct](https://llrs.github.io/experDesign/CODE_OF_CONDUCT.html).
 By participating in this project you agree to abide by its terms.

### 2020-09-24 README.Rmd

@@
 [![Coverage status](https://codecov.io/gh/llrs/experDesign/branch/master/graph/badge.svg)](https://codecov.io/github/llrs/experDesign?branch=master)
-[![lifecycle](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
-[![Project Status: Active - The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
+[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://www.tidyverse.org/lifecycle/#stable)
+[![Project Status: Active - The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active)
 [![R build status](https://github.com/llrs/experDesign/workflows/R-CMD-check/badge.svg)](https://github.com/llrs/experDesign/actions?workflow=R-CMD-check)
@@
 
+To install the latest version on [CRAN](https://CRAN.R-project.org/package=experDesign) use:
+
+```r
+install.packages("experDesign")
+```
 You can install the development version from [GitHub](https://github.com/) with:
@@
 ```
+
 ## Example

### 2020-09-24 README.Rmd
Update Github action to test as CRAN

@@
 <!-- badges: start -->
+[![R build status](https://github.com/llrs/experDesign/workflows/R-CMD-check/badge.svg)](https://github.com/llrs/experDesign/actions?workflow=R-CMD-check)
+[![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/llrs/experDesign?branch=master&svg=true)](https://ci.appveyor.com/project/llrs/experDesign)
 [![Travis build status](https://travis-ci.org/llrs/experDesign.svg?branch=master)](https://travis-ci.org/llrs/experDesign)
-[![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/llrs/experDesign?branch=master&svg=true)](https://ci.appveyor.com/project/llrs/experDesign)
 [![Coverage status](https://codecov.io/gh/llrs/experDesign/branch/master/graph/badge.svg)](https://codecov.io/github/llrs/experDesign?branch=master)
@@
 [![Project Status: Active - The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active)
-[![R build status](https://github.com/llrs/experDesign/workflows/R-CMD-check/badge.svg)](https://github.com/llrs/experDesign/actions?workflow=R-CMD-check)
 <!-- badges: end -->

### 2020-09-24 R/indexing.R
Increase coverage of tests

@@ -35,7 +35,7 @@ create_subset <- function(size_data, size_subset = NULL, n = NULL, name = "SubSe
   if (length(name) != 1 & length(name) != n) {
     stop("Provide a single character or a vector the same size of the batches.",
          call. = FALSE)
-  } else if (length(names) == 1) {
+  } else if (length(name) == 1) {
     names_batches <- paste0(name, seq_len(n))
   } else {
     names_batches <- name

### 2020-09-24 R/utils.R

@@ -18,7 +18,21 @@ omit <- function(pheno, omit){
   }
 }
 
+position_parse <- function(x){
 
+}
+
+
+position_name <- function(rows, columns) {
+  nrow <- length(rows)
+  ncol <- length(columns)
+  plate <- matrix(nrow = nrow, ncol = ncol, dimnames = list(rows, columns))
+  positions <- expand.grid(rows, columns, stringsAsFactors = FALSE)
+  positions$Var2 <- as.character(positions$Var2)
+  positions$name <- apply(positions, 1, paste0, collapse = "")
+  colnames(positions)[1:2] <- c("row", "column")
+  positions
+}
 
 summary_num <- function(pheno) {
   if (is.null(ncol(pheno))) {

### 2020-09-24 tests/testthat/test-distribution.R

@@ -1,0 +1,10 @@
+test_that("distribution works", {
+  data(survey, package = "MASS")
+  set.seed(4565)
+  columns <- c("Sex", "Age", "Smoke")
+  index <- design(pheno = survey[, columns], size_subset = 70,
+                  iterations = 10)
+  batches <- inspect(index, survey[, columns])
+  expect_true(distribution(batches, "Sex"))
+  expect_true(distribution(batches, "Smoke"))
+})

### 2020-09-24 tests/testthat/test-evaluate_entropy.R

@@ -1,0 +1,9 @@
+test_that("evaluate_entropy works", {
+  data(survey, package = "MASS")
+  index <- design(survey[, c("Sex", "Smoke", "Age")], size_subset = 50,
+                  iterations = 50)
+  # Note that numeric columns will be omitted:
+  out <- evaluate_entropy(index, survey[, c("Sex", "Smoke", "Age")])
+  expect_length(out, 2L)
+  expect_equal(names(out), c("Sex", "Smoke"))
+})

### 2020-09-24 tests/testthat/test-inspect.R

@@ -1,0 +1,12 @@
+test_that("inspect works", {
+  data(survey, package = "MASS")
+  columns <- c("Sex", "Age", "Smoke")
+  index <- design(pheno = survey[, columns], size_subset = 70,
+                  iterations = 10)
+  batches <- inspect(index, survey[, columns])
+  expect_true("batch" %in% colnames(batches))
+
+
+  batches <- inspect(index, survey, omit = colnames(survey)[!colnames(survey) %in% columns])
+  expect_true("batch" %in% colnames(batches))
+})

### 2020-09-24 R/spatial.R
Add spatial function

@@ -1,0 +1,87 @@
+#' Distribute the sample on the plate
+#'
+#' This function assumes that to process the batch the samples are distributes in
+#' a plate. Sometimes you know in advance the
+#' @inheritParams design
+#' @param index A list with the samples on each subgroup, as provided from
+#' `design()` or `replicates()`.
+#' @param remove_positions Character, name of positions.
+#' @param rows Character, name of the rows to be used.
+#' @param columns Character, name of the rows to be used.
+#' @return The indices of which samples go with which batch.
+#' @export
+#' @examples
+#' data(survey, package = "MASS")
+#' index <- design(survey[, c("Sex", "Smoke", "Age")], size_subset = 50,
+#'                 iterations = 25)
+#' index2 <- spatial(index, survey[, c("Sex", "Smoke", "Age")], iterations = 25)
+#' head(index2)
+spatial <- function(index, pheno, omit = NULL, remove_positions = NULL, rows = LETTERS[1:5],
+         columns = 1:10, iterations = 500) {
+
+  nrow <- length(rows)
+  ncol <- length(columns)
+
+  if (is.null(rows) | length(rows) == 0) {
+    stop("Please provide at least one row.", call. = FALSE)
+  }
+  if (is.null(columns) | length(columns) == 0) {
+    stop("Please provide at least one column.", call. = FALSE)
+  }
+  if ((nrow*ncol-length(remove_positions)) < max(lengths(index))) {
+    stop("The size for the batch is smaller than the samples it must contain.",
+         "\n\tPlease check the rows and columns or how you created the index.",
+         call. = FALSE)
+  }
+  plate <- matrix(nrow = nrow, ncol = ncol, dimnames = list(rows, columns))
+  positions <- position_name(rows, columns)
+  if (any(!remove_positions %in% positions$name)) {
+    stop("Unrecognized position to remove.",
+         "\n\tCheck that it is a combination of rows and columns: A1, A3...",
+         call. = FALSE)
+  }
+
+  position <- positions$name[!positions$name %in% remove_positions]
+
+  opt <- Inf
+
+  # Calculate batches
+  batches <- length(index)
+  pheno_o <- omit(pheno, omit)
+
+  original_pheno <- evaluate_orig(pheno_o)
+  original_pheno["na", ] <- original_pheno["na", ]/batches
+
+  # Find the numeric values
[…]

### 2020-09-24 README.Rmd
Update README

@@
 <!-- badges: start -->
+[![CRAN status](https://www.r-pkg.org/badges/version/experDesign)](https://CRAN.R-project.org/package=experDesign)
 [![R build status](https://github.com/llrs/experDesign/workflows/R-CMD-check/badge.svg)](https://github.com/llrs/experDesign/actions?workflow=R-CMD-check)

### 2020-09-25 README.Rmd
Change link to https

--- 2020-09-24 README.Rmd
@@
 [![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://www.tidyverse.org/lifecycle/#stable)
-[![Project Status: Active - The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active)
+[![Project Status: Active - The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
 <!-- badges: end -->

### 2020-09-28 README.Rmd
Change OSAT link
Change link to CoC and version

@@
 
-- The [OSAT](https://bioconductor.org/packages/OSAT/) package handles categorical 
+- The [OSAT](https://bioconductor.org/packages/release/bioc/html/OSAT.html) package handles categorical 
 variables but not numeric data. It doesn't work with our data.
@@
 
-Please note that this project is released with a [Contributor Code of Conduct](https://llrs.github.io/experDesign/CODE_OF_CONDUCT.html).
+Please note that this project is released with a [Contributor Code of Conduct](https://www.contributor-covenant.org/version/1/0/0/code-of-conduct/).
 By participating in this project you agree to abide by its terms.

### 2021-04-07 paper.Rmd
Further extension of the text, specially statement of need

---
title: "experDesign: helping performing experiments on collections"
output: 
  md_document:
    preserve_yaml: true
authors:
  - name: "Lluís Revilla Sancho"
    orcid: 0000-0001-9747-2570
    affiliation: '1, 2'
  - name: "Juan José Lozano Salvatella"
    orcid: 0000-0001-9747-2570
    affiliation: '1'
  - name: "Azucena Salas Martínez"
    orcid: 0000-0001-9747-2570
    affiliation: '2'
affiliations:
  - name: "Centro de Investigación Biomédica en Red, Enfermedades Hepáticas y Digestivas"
    index: 1
  - name: "Institut d'Investigacions Biomèdiques August Pi i Sunyer, IDIBAPS"
    index: 2
date: Sys.Date()
tags:
  - R
  - Batch effect
bibliography: paper.bib
---

# Summary

The design of an experiment is critical for their success. However, after a correct design the collection of samples is a critical step where many errors and problems might affect the experiment result. Missing samples on experiments that must be carried out on batches might impact the results if not taken into account. experDesign helps to minimize batch effects for known variables assigning which samples go on which batch.

# Statement of need

On the design of an experiment three main techniques are used to see differences between groups of interest: blocks, then randomization and replication [@klaus2015]. Sometimes variables caused by not tacking into account the blocking effect they have create batch effects on the experiment. There are several techniques to identify and take into account them [@leek2010] . Randomization is a method to get the average effect by mixing the potential confounding variables. Replication is important to judge whether an observed difference is merely by chance or there is an actual difference between the variables. It consists on adding more samples with the same conditions in order to be able to estimate the dispersion and the mean value of the population. On some settings (clínical, agriculture, ...) several of these techniques are applied together to ensure the robustness of the study.

Between the designing of an experiment and the collection of the samples many things can happen. If something goes wrong and some samples are missing they can increase confounding variables or even make the study useless. However, few tools exists to prevent generating a batch effect.

To prevent the batch effect there can be two options: randomization and/or replication. Whether the practitioner goes with randomization or replication they must take into account the blocking design from the beginning or they might. For instance, if cases and controls are run in different batches, sample variation can be entirely confounded by this [@chen2011]. The samples can be properly randomized to minimize the batch effect by looking how the variables distribute on each batch.

The package experDesign for R is an extension to R that all[@rcoreteam2014]

Several functions of experDesign allow this. The main function `design`distribute the samples on multiple batches so that each variable is equally homogeneous in each batch. If the experiment is carried out on a specific spatial distribution the `spatial` function takes into consideration the position of the samples to distribute them homogeneously also by position.

Another recommended quality control measure to avoid batch effects is adding replicates. The most common method is adding technical replicates [@blainey2014]. Technical replicates are samples that are divided on several batches so that later each batch can be compared based on this sample repeated across multiple batches. To make it easier to select the replicates needed and from which samples the function `replicates` is provided.

# References

### 2021-04-09 paper.Rmd
Update paper with some` information about the other packages

@@
 
-# Statement of need
+# Introduction
 
@@
 If something goes wrong and some samples are missing they can increase confounding variables or even make the study useless.
-However, few tools exists to prevent generating a batch effect.
 
@@
 For instance, if cases and controls are run in different batches, sample variation can be entirely confounded by this [@chen2011].
-The samples can be properly randomized to minimize the batch effect by looking how the variables distribute on each batch.
+The samples can be properly randomized to minimize the batch effect by looking how the variables distribute on each batch, on what it is know as randomized blocked/ stratified sampling experimental design.
 
-The package experDesign for R is an extension to R that all[@rcoreteam2014]
+# State of the art
 
-Several functions of experDesign allow this.
+There are some tools to prevent generating a batch effect on the R language from multiple fields and areas, but specially on the biological research [@rcoreteam2014].
+They have some caveats that limit their application to some cases and to our knowledge there hasn't been a comparison of these different methods.
+
+*OSAT*, on [Bioconductor](https://bioconductor.org/packages/OSAT/), first allocates the samples on each batch according to a variable and later shuffles the samples on each batch to randomize the other variables [@yan2012].
+This algorithm relies on categorical variables and cannot use numeric ones such as age or time-related variables unless they are treated as categorical.
+
+*minDiff*, on [github](https://github.com/m-Py/minDiff), and its successor *anticlust*, on [CRAN](https://cran.r-project.org/package=anticlust), [@usingan] divided elements into similar parts, ensuring similarity between groups by enforcing heterogeneity within group.
+Conceptually it is similar to reversing the clustering methods k-means and cluster editing, and as them it must use numeric variables.
+
+Recently, *Omixer*, a new package on [Bioconductor](https://bioconductor.org/packages/Omixer/), has been made available [@sinke2021].
+It tests if the random assignments are homogeneous transforming all variables to numeric and using the Kendall's correlation if there are more than 5 samples, otherwise using the Pearson chi-squared test.
+
+Finally the package *experDesign*, on [CRAN](https://cran.r-project.org/package=experDesign), provides similar functionality conceptually similar to *anticlust* but does accept categorical variables.
+
 The main function `design`distribute the samples on multiple batches so that each variable is equally homogeneous in each batch.

### 2021-04-12 paper.Rmd
Some more progress on the comparison and description of the package

--- 2021-04-07 paper.Rmd
@@
   - Batch effect
+  - experiment design
 bibliography: paper.bib
@@
 
-# Statement of need
+# Introduction
 
-On the design of an experiment three main techniques are used to see differences between groups of interest: blocks, then randomization and replication [@klaus2015].
-Sometimes variables caused by not tacking into account the blocking effect they have create batch effects on the experiment.
-There are several techniques to identify and take into account them [@leek2010] .
-Randomization is a method to get the average effect by mixing the potential confounding variables.
-Replication is important to judge whether an observed difference is merely by chance or there is an actual difference between the variables.
-It consists on adding more samples with the same conditions in order to be able to estimate the dispersion and the mean value of the population.
+Sometimes not tacking into account how some variable are blocking creates batch effects on an experiment.
+There are several techniques to identify and take into account batch effects when analyzing an experiment [@leek2010].
+However, it would be better if the batch effects could be avoided before carrying out the experiment.
+To avoid creating them a good design of the experiment is needed.
+To design an experiment three techniques are used to reduce unwanted variation: blocking, randomization and replication [@klaus2015].
+
+**Blocking** groups samples that are equal according to one or more variables allowing to estimate the variation on the measurement from these variables.
+**Randomization** is a method to get the average effect by mixing the potential confounding variables.
+**Replication** increases the number of samples used on the experiment to better estimate the variation of the experiment.
 On some settings (clínical, agriculture, ...) several of these techniques are applied together to ensure the robustness of the study.
@@
 Between the designing of an experiment and the collection of the samples many things can happen.
-If something goes wrong and some samples are missing they can increase confounding variables or even make the study useless.
-However, few tools exists to prevent generating a batch effect.
+If something goes wrong and some samples are missing the unwanted variation of the experiment will increase.
+Sometimes If the experiments needs to be done in batches and if not taken into account on the design of the experiment it can introduce a batch effect that can be impossible to remove from the data.
 
 To prevent the batch effect there can be two options: randomization and/or replication.
-Whether the practitioner goes with randomization or replication they must take into account the blocking design from the beginning or they might.
-For instance, if cases and controls are run in different batches, sample variation can be entirely confounded by this [@chen2011].
-The samples can be properly randomized to minimize the batch effect by looking how the variables distribute on each batch.
+If the practitioner goes with randomization or replication they must take into account the blocking design from the beginning or they might further exacerbate the problem.
+For instance, if cases and controls are measured in different batches, sample variation can be entirely confounded by this [@chen2011].
+Samples can be properly randomized to minimize the batch effect by looking how the variables distribute on each batch, on what it is known as randomized blocked/ stratified sampling experimental design.
 
-The package experDesign for R is an extension to R that all[@rcoreteam2014]
+# State of the art
 
-Several functions of experDesign allow this.
-The main function `design`distribute the samples on multiple batches so that each variable is equally homogeneous in each batch.
-If the experiment is carried out on a specific spatial distribution the `spatial` function takes into consideration the position of the samples to distribute them homogeneously also by position.
+There are some tools to prevent generating a batch effect on the R language from multiple fields and areas, but specially on the biological research [@rcoreteam2014].
+They have some caveats that limit their application to some cases and to our knowledge there hasn't been a comparison of these different methods.
+Here we describe existing packages briefly:
 
-Another recommended quality control measure to avoid batch effects is adding replicates.
-The most common method is adding technical replicates [@blainey2014].
-Technical replicates are samples that are divided on several batches so that later each batch can be compared based on this sample repeated across multiple batches.
-To make it easier to select the replicates needed and from which samples the function `replicates` is provided.
+-   *OSAT*, on [Bioconductor](https://bioconductor.org/packages/OSAT/), first allocates the samples on each batch according to a variable and later shuffles the samples on each batch to randomize the other variables [@yan2012].
+This algorithm relies on categorical variables and cannot use numeric ones such as age or time-related variables unless they are treated as categorical.
+
+-   *minDiff*, on [github](https://github.com/m-Py/minDiff), and its successor *anticlust*, on [CRAN](https://cran.r-project.org/package=anticlust), [@usingan] divides the samples into similar groups, ensuring similarity by enforcing heterogeneity within group.
+Conceptually it is similar to reversing the clustering methods k-means and cluster editing.
+
+-   Recently, *Omixer*, a new package on [Bioconductor](https://bioconductor.org/packages/Omixer/), has been made available [@sinke2021].
+It tests if the random assignments are homogeneous transforming all variables to numeric and using the Kendall's correlation if there are more than 5 samples, otherwise using the Pearson chi-squared test.
+
+-   Finally the package *experDesign*, on [CRAN](https://cran.r-project.org/package=experDesign), which provides groups with characteristics similar to the whole sample by comparing the random samples with the whole dataset for several statistics.
+
+# Description
+
+The package experDesign provides the function `design` to distribute the samples on multiple batches so that each variable is homogeneous in each batch.
+It is similar to the *anticluster* method for maximum variance.
+If the experiment is carried out on a specific spatial distribution the `spatial` function distributes the samples homogeneously also by position similar to *Omixer*.
+Currently it doesn't allow to mask some positions as *Omixer* .
+
+Adding replicates is a method to have a successful experiment.
+When the design is already done then the replicates are called technical replicates [@blainey2014].
+Technical replicates are samples that are measured multiple times and allow to reduce the uncertainty of the measurement.
+If the technical replicates are distributed on several batches they allow to measure the batch effect.
+To select the replicates needed and from which samples the function `extreme_cases` is provided.
+For easier usage the `replicates` designs an experiment with the number of replicates per batch desired.
+
+experDesign also provides functionality to calculate the right number of batches for your data.
+Instead of sequentially filling the batches it distributes the samples on the minimum number of batches required.
 

### 2021-04-14 R/optimum.R
Faster process to identify batch size
Removes the need for expand.grid so it should work for any reasonable size of data

@@ -90,19 +90,18 @@ internal_batches <- function(size_data, size_subset, batches) {
     return(rep(size_subset, times = batches))
   }
 
+  max_batch_size <- optimum_subset(size_data, batches)
   # Calculate the minimum number of samples per batch
-  remaining <- size_data - size_subset*(batches - 1)
-
-  # Look for the combination of samples that is less reductive
-  s <- seq(from = remaining, to = size_subset)
-  l <- rep(list(s), batches) # List of possibilities for each batch
-  combn_sizes <- expand.grid(l) # All combinations
-  combn_sizes <- as.matrix(combn_sizes)
-  k <- rowSums(combn_sizes) == size_data
-  combn_sizes <- combn_sizes[k, ]
-  r <- apply(combn_sizes, 1, function(x){max(x) - min(x)})
-  final_combn <- combn_sizes[r == min(r), ][1, ]
-  sort(unlist(final_combn, FALSE, FALSE), decreasing = TRUE)
+  remaining <- size_data - max_batch_size*batches
+  # Pre-allocate the max size of each batch
+  out <- rep(max_batch_size, batches)
+  # Calculate how many samples must be removed
+  samples_to_remove_per_batch <- ceiling(abs(remaining)/batches)
+  # Calculate how many batches have less samples
+  batches_to_remove_samples <- abs(remaining)/samples_to_remove_per_batch
+  # Apply it:``
+  out[1:batches_to_remove_samples] <- out[1:batches_to_remove_samples] - samples_to_remove_per_batch
+  return(sort(out, decreasing = TRUE))
 }
 
 check_number <- function(x, name) {

### 2021-04-14 tests/testthat/test-spatial.R
Correct test to new size

@@ -6,9 +6,9 @@ test_that("spatial works", {
   index2 <- spatial(index, survey[, c("Sex", "Smoke", "Age")],
                     rows = LETTERS[1:9],
                     columns = 1:12, iterations = 25)
-  expect_length(index2, 50)
-  expect_equivalent(table(lengths(index2)), as.table(c("4" = 13, "5" = 37)))
-  expect_equal(names(index2), position_name(rows = LETTERS[1:5], 1:10)$name)
+  expect_length(index2, 9*12)
+  expect_equivalent(table(lengths(index2)), as.table(c("2" = 87, "3" = 21)))
+  expect_equal(names(index2), position_name(rows = LETTERS[1:9], 1:12)$name)
 })
 
 test_that("spatial works with replicates", {

### 2021-04-14 README.Rmd

@@
 
+ - The [Omixer](https://bioconductor.org/packages/Omixer/) package handles both 
+numeric and categorical data (converting categorical variables to numeric).
+But both the same way either Pearson's Chi-squared Test if there are few samples or Kendall's correlation.
+It does allow to protect some spots from being used.
+
 If you are still designing the experiment and do not have collected any data [DeclareDesign](https://cran.r-project.org/package=DeclareDesign) might be relevant for you.

### 2021-04-14 paper.Rmd
Minor modifications

@@
 If the experiment is carried out on a specific spatial distribution the `spatial` function distributes the samples homogeneously also by position similar to *Omixer*.
-Currently it doesn't allow to mask some positions as *Omixer* .
 
@@
 If the technical replicates are distributed on several batches they allow to measure the batch effect.
-To select the replicates needed and from which samples the function `extreme_cases` is provided.
+To select the technical replicates needed and from which samples the function `extreme_cases` is provided.
 For easier usage the `replicates` designs an experiment with the number of replicates per batch desired.
 
-experDesign also provides functionality to calculate the right number of batches for your data.
-Instead of sequentially filling the batches it distributes the samples on the minimum number of batches required.
+experDesign also provides several small utilities to make it easier design the experiment in batches.
+For instance, it helps to calculate the samples to place on the minimum number of batches required with `sizes_batches`.
 

### 2021-04-14 paper.Rmd
Update content

@@
 ---
-title: "experDesign: helping performing experiments on collections"
+title: "experDesign: helping performing experiments on batches"
 output: 
@@
 Missing samples on experiments that must be carried out on batches might impact the results if not taken into account.
-experDesign helps to minimize batch effects for known variables assigning which samples go on which batch.
+*experDesign* helps to minimize batch effects for known variables assigning which samples go on which batch.
 
@@
 **Replication** increases the number of samples used on the experiment to better estimate the variation of the experiment.
-On some settings (clínical, agriculture, ...) several of these techniques are applied together to ensure the robustness of the study.
+On some settings (clinical, agriculture, ...) several of these techniques are applied together to ensure the robustness of the study.
 
@@
 
-The package experDesign provides the function `design` to distribute the samples on multiple batches so that each variable is homogeneous in each batch.
+The package *experDesign* provides the function `design` to distribute the samples on multiple batches so that each variable is homogeneous in each batch.
 It is similar to the *anticluster* method for maximum variance.
@@
 
-Adding replicates is a method to have a successful experiment.
+In addition to distribute the samples on batches *experDesign* provides with tools to add technical replicates.
 When the design is already done then the replicates are called technical replicates [@blainey2014].
 Technical replicates are samples that are measured multiple times and allow to reduce the uncertainty of the measurement.
-If the technical replicates are distributed on several batches they allow to measure the batch effect.
+If the technical replicates are distributed on several batches they allow to measure the batch effect and thus minimize it.
 To select the technical replicates needed and from which samples the function `extreme_cases` is provided.
@@
 
-experDesign also provides several small utilities to make it easier design the experiment in batches.
-For instance, it helps to calculate the samples to place on the minimum number of batches required with `sizes_batches`.
+*experDesign* also provides several small utilities to make it easier design the experiment in batches.
+For instance, it helps to calculate the samples to dictribute on the minimum number of batches required with `sizes_batches`.
 

### 2021-04-14 paper.Rmd
Fix typo
Fix references

@@
 
--   *minDiff*, on [github](https://github.com/m-Py/minDiff), and its successor *anticlust*, on [CRAN](https://cran.r-project.org/package=anticlust), [@usingan] divides the samples into similar groups, ensuring similarity by enforcing heterogeneity within group.
+-   *minDiff*, on [github](https://github.com/m-Py/minDiff), and its successor *anticlust*, on [CRAN](https://cran.r-project.org/package=anticlust), divides the samples into similar groups, ensuring similarity by enforcing heterogeneity within group [@papenberg].
     Conceptually it is similar to reversing the clustering methods k-means and cluster editing.
@@
 *experDesign* also provides several small utilities to make it easier design the experiment in batches.
-For instance, it helps to calculate the samples to dictribute on the minimum number of batches required with `sizes_batches`.
+For instance, it helps to calculate the samples to distribute on the minimum number of batches required with `sizes_batches`.
 

### 2021-04-14 paper.Rmd
Save changes reorganizing paper

@@
 The design of an experiment is critical for their success.
-However, after a correct design the collection of samples is a critical step where many errors and problems might affect the experiment result.
-Missing samples on experiments that must be carried out on batches might impact the results if not taken into account.
-*experDesign* helps to minimize batch effects for known variables assigning which samples go on which batch.
+However, after an initial correct design the process till the measurement is critical.
+At the several steps required from collection to measurement many errors and problems might affect the experiment result.
+This variation if not taken into account can make an experiment inconclusive.
+*experDesign* provides tools to minimize the risk of making an experiment inconclusive by assigning which samples go on which batch.
 
@@
 
-Sometimes not tacking into account how some variable are blocking creates batch effects on an experiment.
-There are several techniques to identify and take into account batch effects when analyzing an experiment [@leek2010].
-However, it would be better if the batch effects could be avoided before carrying out the experiment.
-To avoid creating them a good design of the experiment is needed.
-To design an experiment three techniques are used to reduce unwanted variation: blocking, randomization and replication [@klaus2015].
+To properly design an experiment the source of variation between the samples must be the one we are interested in.
+Usually one controls the environment where the study or experiment is being done.
+Sometimes, this is not enough or not possible, and then one need to apply some methods to control these variation.
+There are three techniques used to reduce unwanted variation: blocking, randomization and replication [@klaus2015].
 
@@
 
-Between the designing of an experiment and the collection of the samples many things can happen.
-If something goes wrong and some samples are missing the unwanted variation of the experiment will increase.
-Sometimes If the experiments needs to be done in batches and if not taken into account on the design of the experiment it can introduce a batch effect that can be impossible to remove from the data.
+Between the designing of an experiment and the measurement of the samples many things can happen.
+If some samples go missing (or are contaminated, or do not pass the quality control to perform the measurement, or ...) the unwanted variation of the experiment will increase.
+Even if this doesn't happen, sometimes the experiments need to be done in batches because there are some technical; for example the machine cannot measure more samples at once, or practical considerations; for example it is not possible to take more measurements on the field in the allowed time.
 
-To prevent the batch effect there can be two options: randomization and/or replication.
-If the practitioner goes with randomization or replication they must take into account the blocking design from the beginning or they might further exacerbate the problem.
-For instance, if cases and controls are measured in different batches, sample variation can be entirely confounded by this [@chen2011].
-Samples can be properly randomized to minimize the batch effect by looking how the variables distribute on each batch, on what it is known as randomized blocked/ stratified sampling experimental design.
+There are several techniques to identify and take into account batch effects when analyzing an experiment already measured [@leek2010].
+However, if the source of them are not taken into account before measuring they can introduce a batch effect that could be impossible to remove.
+Therefore, it would be better if the batch effects were avoided before carrying out the experiment.
+Tacking into account how the process from the design to the measurement can introduce variability will help avoiding batch effects.
+
+To prevent the apparition of batch effect after the design of the experiment there are two options: randomization and replication.
+This techniques can prevent the batch effect to be able to compare between or within the desired variable.
+One can do both of them but in any case it must take into account the blocking design from the beginning or they might further exacerbate the problem.
+For instance, if one designs an experiment with cases and controls and the controls are measured in one batch and the controls in a different batch then the difference between them can be entirely confounded by a malfunction of the measurement machine in one batch [@chen2011].
+
+However, samples can be properly randomized to minimize the batch effect by looking how the variables distribute on each batch, this is known as randomized block experimental design or stratified random sampling experimental design.
+
+Replications consist on increasing the number of measurements with similar attributes.
+Usually at least three samples are included for each condition of interest to be able to estimate the variation of the attributes.
+If there is one extraction and then a sample is measured multiple times it is called a technical replicate.
+Technical replicates help estimate the variation of the measurement method or the process.
 
@@
 
+# Comparison
+
+For completeness a description of the usage of the different software
+
+```{r example-data}
+data(survey, package = "MASS")
+survey$ID <- seq_len(nrow(survey))
+VoI <- c("Sex", "Smoke", "Age")
+n_batch <- 3
+iterations <- 1000
+```
+
+## OSAT
+
+OSAT provides some template for some setups:
+
+```{r osat}
+library("OSAT")
+gs <- setup.sample(survey, optimal = VoI)
+gc <- setup.container(IlluminaBeadChip96Plate, n_batch, batch = 'plates')
+gSetup <- create.optimized.setup(sample = gs, container = gc, nSim = iterations)
+```
+
+## anticlust
+
+```{r anticlust}
+library("anticlust")
+anticlust_index <- anticlustering(
+  survey[, VoI],
+  K = n_batch,
+  objective = "variance",
+  method = "exchange",
+  repetitions = iterations
+)
+```
+
+## Omixer
+
+```{r Omixer}
+library("Omixer")
+layout <- data.frame(well = 1:96, 
+    row = factor(rep(1:8, each=12), labels = LETTERS[1:8]),
+    column = rep(1:12, 8), 
+    plate  = 1,
+    chip = as.integer(ceiling(rep(1:12, 8)/2)))
+layout$chipPos <- ifelse(layout$column %% 2 == 0, 
+                         as.numeric(layout$row) + 8,
+                         layout$row)
+layout2 <- layout
+layout2$plate <- 2
+layout3 <- layout
+layout3$plate <- 3
+plates <- rbind(layout, layout2, layout3)
+Omixer_index <- omixerRand(survey, sampleId = "ID", 
+    block = "block", iterNum = 10, wells = 96,
+    plateNum = n_batch, randVars = VoI, layout = plates, techVars = NULL)
+```
+
+## experDesign
+
+```{r experDesign}
+library("experDesign")
+experDesign_index <- design(survey[, VoI], size_subset = 96,
+                  iterations = iterations)
+```
+
 # References

### 2021-04-14 tests/testthat/test-optimum.R
Returns the maximum number of samples possible or the just required number of

@@ -23,4 +23,10 @@ test_that("internal_batches", {
   batches <- 1
   expect_length(internal_batches(size_data, size_batch, batches), batches)
   expect_true(internal_batches(size_data, size_batch, batches) <= size_batch)
+  out <- internal_batches(237, 96, 3)
+  expect_equal(out, c(79, 79, 79))
+  out <- internal_batches(237, 79, 3)
+  expect_equal(out, c(79, 79, 79))
+  out <- internal_batches(237, 78, 3)
+  expect_equal(out, c(78, 78, 78))
 })

### 2021-04-15 paper.Rmd
More tweaks

@@
 gs <- setup.sample(survey, optimal = VoI)
-gc <- setup.container(IlluminaBeadChip96Plate, n_batch, batch = 'plates')
+gc <- setup.container(IlluminaBeadChip96Plate, n = n_batch, batch = 'plates')
 gSetup <- create.optimized.setup(sample = gs, container = gc, nSim = iterations)
@@
   survey[, VoI],
+  K = n_batch,
+  objective = "variance",
+  method = "exchange",
+  repetitions = iterations
+)
+anticlust_index <- anticlustering(
+  survey[, "Age"],
   K = n_batch,
@@
 
+## Summary
+
+Last we can compare some of the solutions
+
+```{r compare}
+OSAT_index <- get.experiment.setup(gSetup)$plates
+table(batch_names(experDesign_index), anticlust_index)
+# OSAT omits one sample
+table(batch_names(experDesign_index)[-137], OSAT_index$plates)
+table(anticlust_index[-137], OSAT_index$plates)
+```
+
 # References

### 2021-04-16 paper.Rmd
Some more tweaks

--- 2021-04-14 paper.Rmd
@@
 
-For completeness a description of the usage of the different software
+For completeness a brief examples using each different software.
+
+First of all a brief data setup.
 
 ```{r example-data}
-data(survey, package = "MASS")
-survey$ID <- seq_len(nrow(survey))
-VoI <- c("Sex", "Smoke", "Age")
-n_batch <- 3
-iterations <- 1000
+data(cats, package = "MASS")
+cats$ID <- seq_len(nrow(cats))
+VoI <- c("Sex", "Bwt", "Hwt")
+n_batch <- 2
+iterations <- 10000
 ```
+
+The objective is to group the cats in 2 groups to avoid any batch effect on further measures on them.
+We set 10000 iterations and 2 groups
 
@@
 
-OSAT provides some template for some setups:
+OSAT provides some template for some layouts:
 
@@
 library("OSAT")
-gs <- setup.sample(survey, optimal = VoI)
-gc <- setup.container(IlluminaBeadChip96Plate, n_batch, batch = 'plates')
+gs <- setup.sample(cats, optimal = VoI)
+gc <- setup.container(IlluminaBeadChip96Plate, n = n_batch, batch = 'plates')
 gSetup <- create.optimized.setup(sample = gs, container = gc, nSim = iterations)
@@
 
+anticlust only works with numeric variables:
+
 ```{r anticlust}
@@
 anticlust_index <- anticlustering(
-  survey[, VoI],
+  cats[, VoI],
+  K = n_batch,
+  objective = "variance",
+  method = "exchange",
+  repetitions = iterations
+)
+anticlust_index <- anticlustering(
+  cats[, c("Bwt", "Hwt")],
   K = n_batch,
@@
 layout2$plate <- 2
-layout3 <- layout
-layout3$plate <- 3
-plates <- rbind(layout, layout2, layout3)
-Omixer_index <- omixerRand(survey, sampleId = "ID", 
-    block = "block", iterNum = 10, wells = 96,
+plates <- rbind(layout, layout2)
+Omixer_index <- omixerRand(cats, sampleId = "ID", 
+    block = "none", iterNum = iterations, wells = 96,
     plateNum = n_batch, randVars = VoI, layout = plates, techVars = NULL)
@@
 library("experDesign")
-experDesign_index <- design(survey[, VoI], size_subset = 96,
+experDesign_index <- design(cats[, VoI], size_subset = 96,
                   iterations = iterations)
@@
 
+## Summary
+
+Last we can compare some of the solutions
+
+```{r compare}
+OSAT_index <- get.experiment.setup(gSetup)$plates
+table(batch_names(experDesign_index), anticlust_index)
+
+table(batch_names(experDesign_index), OSAT_index)
+table(anticlust_index, OSAT_index)
+```
+
+It doesn't seem possible to run Omixer without any spatial batch effect
+
 # References

### 2021-04-20 paper.Rmd
Update with code

@@
 ---
+
+```{r setup}
+knitr::opts_chunk$set(collapse = TRUE, warning = TRUE, error = TRUE)
+```
 
@@
 If the technical replicates are distributed on several batches they allow to measure the batch effect and thus minimize it.
-To select the technical replicates needed and from which samples the function `extreme_cases` is provided.
+To select the technical replicates needed and choose from which samples they are done the function `extreme_cases` is provided.
 For easier usage the `replicates` designs an experiment with the number of replicates per batch desired.

### 2021-04-21 paper.Rmd
Remove setup from the output
Add word document, update authors name, remove code

--- 2021-04-15 paper.Rmd
@@
 output: 
+  word_document: 
+    keep_md: true
   md_document:
@@
     affiliation: '1, 2'
-  - name: "Juan José Lozano Salvatella"
+  - name: "Juan-José Lozano"
     orcid: 0000-0001-9747-2570
     affiliation: '1'
-  - name: "Azucena Salas Martínez"
+  - name: "Azucena Salas"
     orcid: 0000-0001-9747-2570
@@
   - R
-  - Batch effect
+  - batch effect
   - experiment design
@@
 There are several techniques to identify and take into account batch effects when analyzing an experiment already measured [@leek2010].
+Experiments that run over long periods of time or are run across different laboratories are highly likely to be susceptible and even smaller single-laboratory studies if spanned several days or include personnel changes may be affected.
 However, if the source of them are not taken into account before measuring they can introduce a batch effect that could be impossible to remove.
 Therefore, it would be better if the batch effects were avoided before carrying out the experiment.
-Tacking into account how the process from the design to the measurement can introduce variability will help avoiding batch effects.
+Tacking into account the process from the design to the measurement can help to avoid batch effects.
 
-To prevent the apparition of batch effect after the design of the experiment there are two options: randomization and replication.
-This techniques can prevent the batch effect to be able to compare between or within the desired variable.
+Experiments should be designed to distribute batches and other potential sources of experimental variation across groups.
+To prevent the apparition of batch effect after the initial design of the experiment there are two options: randomization and replication.
+This techniques can prevent the batch effect and allow to compare between or within the desired variable(s).
 One can do both of them but in any case it must take into account the blocking design from the beginning or they might further exacerbate the problem.
+
+Randomization if done correctly helps reducing the variation across groups while replication let us be more confident of the result of the experiment.
 For instance, if one designs an experiment with cases and controls and the controls are measured in one batch and the controls in a different batch then the difference between them can be entirely confounded by a malfunction of the measurement machine in one batch [@chen2011].
-
-However, samples can be properly randomized to minimize the batch effect by looking how the variables distribute on each batch, this is known as randomized block experimental design or stratified random sampling experimental design.
+However, by looking how the variables distribute on each batch samples can be properly randomized to minimize the batch effect, this is known as randomized block experimental design or stratified random sampling experimental design.
 
@@
 If there is one extraction and then a sample is measured multiple times it is called a technical replicate.
-Technical replicates help estimate the variation of the measurement method or the process.
+Technical replicates help estimate the variation of the measurement method and thus of the possible batch effect.
 
@@
 They have some caveats that limit their application to some cases and to our knowledge there hasn't been a comparison of these different methods.
-Here we describe existing packages briefly:
+Here we describe the current packages briefly:
 
@@
 -   Recently, *Omixer*, a new package on [Bioconductor](https://bioconductor.org/packages/Omixer/), has been made available [@sinke2021].
-    It tests if the random assignments are homogeneous transforming all variables to numeric and using the Kendall's correlation if there are more than 5 samples, otherwise using the Pearson chi-squared test.
+    It tests if the random assignments are homogeneous transforming all variables to numeric and using the Kendall's correlation if there are more than 5 samples, otherwise using Pearson's chi-squared test.
 
@@
 
-The package *experDesign* provides the function `design` to distribute the samples on multiple batches so that each variable is homogeneous in each batch.
+The package ***experDesign*** provides the function `design` to distribute the samples on multiple batches so that each variable is homogeneous in each batch.
 It is similar to the *anticluster* method for maximum variance.
@@
 
-In addition to distribute the samples on batches *experDesign* provides with tools to add technical replicates.
+In addition to distribute the samples on batches ***experDesign*** provides with tools to add technical replicates.
 When the design is already done then the replicates are called technical replicates [@blainey2014].
@@
 If the technical replicates are distributed on several batches they allow to measure the batch effect and thus minimize it.
-To select the technical replicates needed and from which samples the function `extreme_cases` is provided.
+To select the technical replicates needed and choose from which samples they are done the function `extreme_cases` is provided.
 For easier usage the `replicates` designs an experiment with the number of replicates per batch desired.
 
-*experDesign* also provides several small utilities to make it easier design the experiment in batches.
+***experDesign*** also provides several small utilities to make it easier design the experiment in batches.
 For instance, it helps to calculate the samples to distribute on the minimum number of batches required with `sizes_batches`.
 
-# Comparison
-
-For completeness a description of the usage of the different software
-
-```{r example-data}
-data(survey, package = "MASS")
-survey$ID <- seq_len(nrow(survey))
-VoI <- c("Sex", "Smoke", "Age")
-n_batch <- 3
-iterations <- 1000
-```
-
-## OSAT
-
-OSAT provides some template for some setups:
-
-```{r osat}
-library("OSAT")
-gs <- setup.sample(survey, optimal = VoI)
-gc <- setup.container(IlluminaBeadChip96Plate, n = n_batch, batch = 'plates')
-gSetup <- create.optimized.setup(sample = gs, container = gc, nSim = iterations)
-```
-
-## anticlust
-
-```{r anticlust}
-library("anticlust")
-anticlust_index <- anticlustering(
-  survey[, VoI],
-  K = n_batch,
-  objective = "variance",
-  method = "exchange",
-  repetitions = iterations
-)
-anticlust_index <- anticlustering(
-  survey[, "Age"],
-  K = n_batch,
-  objective = "variance",
-  method = "exchange",
-  repetitions = iterations
-)
-```
-
-## Omixer
-
-```{r Omixer}
-library("Omixer")
-layout <- data.frame(well = 1:96, 
-    row = factor(rep(1:8, each=12), labels = LETTERS[1:8]),
-    column = rep(1:12, 8), 
-    plate  = 1,
-    chip = as.integer(ceiling(rep(1:12, 8)/2)))
-layout$chipPos <- ifelse(layout$column %% 2 == 0, 
-                         as.numeric(layout$row) + 8,
-                         layout$row)
-layout2 <- layout
-layout2$plate <- 2
-layout3 <- layout
-layout3$plate <- 3
-plates <- rbind(layout, layout2, layout3)
-Omixer_index <- omixerRand(survey, sampleId = "ID", 
-    block = "block", iterNum = 10, wells = 96,
-    plateNum = n_batch, randVars = VoI, layout = plates, techVars = NULL)
-```
-
-## experDesign
-
-```{r experDesign}
-library("experDesign")
-experDesign_index <- design(survey[, VoI], size_subset = 96,
-                  iterations = iterations)
-```
-
-## Summary
-
-Last we can compare some of the solutions
-
-```{r compare}
-OSAT_index <- get.experiment.setup(gSetup)$plates
-table(batch_names(experDesign_index), anticlust_index)
-# OSAT omits one sample
-table(batch_names(experDesign_index)[-137], OSAT_index$plates)
-table(anticlust_index[-137], OSAT_index$plates)
-```
-
 # References

### 2021-04-21 tests/testthat/test-inspect.R
Improve reporting for replicates

@@ -10,3 +10,19 @@ test_that("inspect works", {
   batches <- inspect(index, survey, omit = colnames(survey)[!colnames(survey) %in% columns])
   expect_true("batch" %in% colnames(batches))
 })
+
+test_that("inspect works with replicates", {
+  data(survey, package = "MASS")
+  columns <- c("Sex", "Age", "Smoke")
+  index <- replicates(pheno = survey[, columns], size_subset = 70, controls = 5,
+                  iterations = 10)
+  survey$sample <- seq_len(nrow(survey))
+  batches <- inspect(index, survey)
+  t_samples <- table(batches$sample)
+  expect_equal(table(t_samples), structure(c(`1` = 232L, `4` = 5L), .Dim = 2L, .Dimnames = list(
+    t_samples = c("1", "4")), class = "table"))
+
+  # TODO check order is right for the samples
+  k <- batches$sample %in% names(t_samples[t_samples == 4])
+  expect_true(all(table(batches$batch[k], batches$sample[k]) <= 1))
+})

### 2021-04-21 README.Rmd
Update README and NEWS

@@
 [![Coverage status](https://codecov.io/gh/llrs/experDesign/branch/master/graph/badge.svg)](https://codecov.io/github/llrs/experDesign?branch=master)
-[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://www.tidyverse.org/lifecycle/#stable)
+[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)
 [![Project Status: Active - The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
@@
 The goal of experDesign is to help you decide which samples go in which batch, 
-reducing the potential batch bias when analyzing.
+reducing the potential batch bias before performing an experiment.
+It provides three main functions :
+
+* `design()`: Randomize the samples according to their variables.
+* `replicates()`: Selects some samples for replicates and randomizes the samples.
+* `spatial()`: Randomize the samples on a spatial grid.
 
@@
 
-Imagine you have some samples already collected and you want to distributed them in batches:
-```{r experDesign_setup}
+We can use the survey dataset for the examples:
+
+```{r show}
 library("experDesign")
-metadata <- expand.grid(height = seq(60, 80, 5), 
-                        weight = seq(100, 300, 50),
-                        sex = c("Male","Female"))
-head(metadata, 15)
+data(survey, package = "MASS") 
+head(survey)
 ```
 
-If you block incorrectly and end up with a group in a single batch we will end up with batch effect.
-In order to avoid this `design` helps you assign each sample to a batch (in this case each batch has 24 samples at most). First we can explore the number of samples and the number of batches:
+The dataset has numeric, categorical values and some `NA`'s value.
 
-```{r size}
-size_data <- nrow(metadata)
-size_batch <- 24
-(batches <- optimum_batches(size_data, size_batch))
-# So now the best number of samples for each batch is less than the available
-(size <- optimum_subset(size_data, batches))
-# The distribution of samples per batch
-sizes_batches(size_data, size, batches)
+# Picking samples for each batch
+
+Imagine that we can only work in groups of 70, and we want to randomize by Sex, 
+Smoke, Age, and by writing hand.  
+There are `r choose(237, 70)` combinations some of them would be have in a 
+single experiment all the right handed students. We could measure all these combinations
+but we can try to find an optimum value.
+
+```{r design, fig.show='hold'}
+# To reduce the variables used:
+omit <- c("Wr.Hnd", "NW.Hnd", "Fold", "Pulse", "Clap", "Exer", "Height", "M.I")
+(keep <- colnames(survey)[!colnames(survey) %in% omit])
+head(survey[, keep])
+
+# Looking for groups at most of 70 samples.
+index <- design(pheno = survey, size_subset = 70, omit = omit)
+index
 ```
 
-Note that instead of using a whole batch and then leave a single sample on the third distributes all the samples in the three batches that will be needed.
-We can directly look for the distribution of the samples given our max number of samples per batch:
+We can transform then into a vector to append to the file or to pass to the lab mate with:
 
-```{r design}
-d <- design(metadata, size_batch)
-# It is a list but we can convert it to a vector with:
-batch_names(d)
+```{r batch_names}
+head(batch_names(index))
 ```
-
-Naively one would either fill some batches fully or distribute them not evenly 
-(the  first 17 packages together, the next 17 and so on). This solution ensures 
-that the data is randomized. For more random distribution you can increase the number of iterations performed to calculate this distribution.
-
-If you need space for replicates to control for batch effect you can use:
-
-```{r replicates}
-r <- replicates(metadata, size_batch, 5)
-lengths(r)
-r
-```
-
-Which seeks as controls the most diverse values and adds them to the samples 
-distribution. Note that if the sample is already present on that batch is not added again, that's why the number of samples per batch is different from the design without replicates.
 

### 2021-04-21 paper.Rmd
Update with revisions & fix date

@@
 ---
-title: "experDesign: helping performing experiments on batches"
+title: 'experDesign: helping performing experiments on batches'
 output:
+  md_document:
+    preserve_yaml: yes
+  html_document:
+    df_print: paged
   word_document:
-    keep_md: true
-  md_document:
-    preserve_yaml: true
+    keep_md: yes
+date: "23 April 2021"
+affiliations:
+- name: Centro de Investigación Biomédica en Red, Enfermedades Hepáticas y Digestivas
+  index: 1
+- name: Institut d'Investigacions Biomèdiques August Pi i Sunyer, IDIBAPS
+  index: 2
 authors:
-  - name: "Lluís Revilla Sancho"
+- name: Lluís Revilla Sancho
   orcid: 0000-0001-9747-2570
-    affiliation: '1, 2'
-  - name: "Juan-José Lozano"
+  affiliation: 1, 2
+- name: Juan-José Lozano
   orcid: 0000-0001-9747-2570
   affiliation: '1'
-  - name: "Azucena Salas"
+- name: Azucena Salas
   orcid: 0000-0001-9747-2570
   affiliation: '2'
-affiliations:
-  - name: "Centro de Investigación Biomédica en Red, Enfermedades Hepáticas y Digestivas"
-    index: 1
-  - name: "Institut d'Investigacions Biomèdiques August Pi i Sunyer, IDIBAPS"
-    index: 2
-date: Sys.Date()
 tags:
@@
 
-The design of an experiment is critical for their success.
-However, after an initial correct design the process till the measurement is critical.
-At the several steps required from collection to measurement many errors and problems might affect the experiment result.
-This variation if not taken into account can make an experiment inconclusive.
-*experDesign* provides tools to minimize the risk of making an experiment inconclusive by assigning which samples go on which batch.
+The design of an experiment is critical to its success.
+Nonetheless, even with a correct design, the process up to the moment of measurement is critical.
+At any one of the several steps required from collection to measurement various errors and problems could affect the experimental results.
+Failure to take such variability into account can render an experiment inconclusive.
+*experDesign* provides tools to minimize the risk of inconclusive results by specifically assigning samples to particular batches.
 
@@
 
-To properly design an experiment the source of variation between the samples must be the one we are interested in.
-Usually one controls the environment where the study or experiment is being done.
-Sometimes, this is not enough or not possible, and then one need to apply some methods to control these variation.
+To properly design an experiment, the source of the variation between samples must be identified.
+Typically, one can control the environment in which the study or experiment is being conducted.
+Sometimes, however, this is not possible, and then one needs to apply certain techniques to control for such variations.
 There are three techniques used to reduce unwanted variation: blocking, randomization and replication [@klaus2015].
 
-**Blocking** groups samples that are equal according to one or more variables allowing to estimate the variation on the measurement from these variables.
-**Randomization** is a method to get the average effect by mixing the potential confounding variables.
-**Replication** increases the number of samples used on the experiment to better estimate the variation of the experiment.
-On some settings (clinical, agriculture, ...) several of these techniques are applied together to ensure the robustness of the study.
+**Blocking** group samples that are equal, according to one or more variables, allows one to estimate the variation in the measurements of the factors from these variables.
+**Randomization** is a method to minimize the variation in the measurements by mixing the potential confounding variables.
+**Replication** increases the number of samples used in an experiment to better estimate the variation of the experiment.
+In some settings (e.g., clinical, agriculture, etc.), several of these techniques can be applied together to ensure the robustness of the study.
 
-Between the designing of an experiment and the measurement of the samples many things can happen.
-If some samples go missing (or are contaminated, or do not pass the quality control to perform the measurement, or ...) the unwanted variation of the experiment will increase.
-Even if this doesn't happen, sometimes the experiments need to be done in batches because there are some technical; for example the machine cannot measure more samples at once, or practical considerations; for example it is not possible to take more measurements on the field in the allowed time.
+Between the designing of an experiment and the measurement of the samples, many things can happen.
+If some samples are lost, are contaminated, or do not pass quality control for measurement purposes, the unwanted variation of the experiment will increase.
+Even if this doesn't happen, experiments will occasionally need to be carried out in batches due to technical reasons; for example, the machine cannot measure more than one samples at a time, or because of other practical considerations; for instance, it may not be possible to obtain additional measurements in the field during the allotted time.
 
-There are several techniques to identify and take into account batch effects when analyzing an experiment already measured [@leek2010].
-Experiments that run over long periods of time or are run across different laboratories are highly likely to be susceptible and even smaller single-laboratory studies if spanned several days or include personnel changes may be affected.
-However, if the source of them are not taken into account before measuring they can introduce a batch effect that could be impossible to remove.
-Therefore, it would be better if the batch effects were avoided before carrying out the experiment.
-Tacking into account the process from the design to the measurement can help to avoid batch effects.
+There are several techniques to identify and assess batch effects when analyzing an already measured experiment [@leek2010].
+Experiments that run over long periods of time or that are run across different laboratories are highly susceptible to batch effects.
+This is also true of even smaller single-laboratory studies, if they span several days or include personnel changes.
+However, if the source of variations are not taken into account before measuring is carried out, a batch effect can be introduced that later is almost impossible to remove.
+Thus, it would be better to avoid such batch effects before carrying out an experiment.
+Taking into account the process from the design phase to the final measurement can help prevent batch effects.
 
 Experiments should be designed to distribute batches and other potential sources of experimental variation across groups.
-To prevent the apparition of batch effect after the initial design of the experiment there are two options: randomization and replication.
-This techniques can prevent the batch effect and allow to compare between or within the desired variable(s).
-One can do both of them but in any case it must take into account the blocking design from the beginning or they might further exacerbate the problem.
+To prevent the occurrence of batch effects after the initial design of the experiment, there are two options: randomization and replication.
+These techniques can not only prevent batch effects, but can also facilitate comparisons between or within the desired variable(s).
+Both can be utilized, but in any case one must take blocking into account from the beginning or the problem might be further exacerbated.
 
-Randomization if done correctly helps reducing the variation across groups while replication let us be more confident of the result of the experiment.
-For instance, if one designs an experiment with cases and controls and the controls are measured in one batch and the controls in a different batch then the difference between them can be entirely confounded by a malfunction of the measurement machine in one batch [@chen2011].
-However, by looking how the variables distribute on each batch samples can be properly randomized to minimize the batch effect, this is known as randomized block experimental design or stratified random sampling experimental design.
+Randomization, if done correctly, can helps reducing variations across groups, while replication can help one be more confident of the experimental results.
+For instance, if one designs an experiment with cases and controls and the latter are measured in one batch and the controls in a different batch, then any difference between them can be entirely confounded by a malfunction of the measuring machine in either one of the batches [@chen2011].
+However, by examining how the variables are distributed across each batch sample, proper randomized can be ensured, thus minimizing batch effects.
+This is known as randomized block experimental design or stratified random sampling experimental design.
+
+Replications consist of increasing the number of measurements with similar attributes.
+Usually at least three samples are included for each condition of interest in order to estimate the variations of these attributes.
+When there is one extraction and then a sample is measured multiple times, this is referred to as a technical replicate.
+Technical replicates help estimate the variation of the measurement method, and thus the possible batch effect.
 
@@
 
-There are some tools to prevent generating a batch effect on the R language from multiple fields and areas, but specially on the biological research [@rcoreteam2014].
-They have some caveats that limit their application to some cases and to our knowledge there hasn't been a comparison of these different methods.
-Here we describe the current packages briefly:
+There are some tools to prevent batch effects on the R language in multiple fields and areas, and particularly for biological research [@rcoreteam2014].
+They carry some caveats limiting their application in some cases, and to our knowledge no comparisons of these various methods has been conducted.
+Here we briefly describe the currently available packages:
 
--   *OSAT*, on [Bioconductor](https://bioconductor.org/packages/OSAT/), first allocates the samples on each batch according to a variable and later shuffles the samples on each batch to randomize the other variables [@yan2012].
-    This algorithm relies on categorical variables and cannot use numeric ones such as age or time-related variables unless they are treated as categorical.
+-   *OSAT*, at [Bioconductor](https://bioconductor.org/packages/OSAT/), first allocates the samples from each batch according to a variable; it then shuffles the samples from each batch in order to randomize the other variables [@yan2012].
+    This algorithm relies on categorical variables and cannot use numerical variables (e.g., age- or time-related) unless they are treated as categorical.
 
--   *minDiff*, on [github](https://github.com/m-Py/minDiff), and its successor *anticlust*, on [CRAN](https://cran.r-project.org/package=anticlust), divides the samples into similar groups, ensuring similarity by enforcing heterogeneity within group [@papenberg].
-    Conceptually it is similar to reversing the clustering methods k-means and cluster editing.
+-   *minDiff*, at [github](https://github.com/m-Py/minDiff), and its successor *anticlust*, at [CRAN](https://cran.r-project.org/package=anticlust), divide the samples into similar groups, ensuring similarity by enforcing heterogeneity within groups [@papenberg2020].
+    Conceptually it is similar to the clustering methods k-means.
 
--   Recently, *Omixer*, a new package on [Bioconductor](https://bioconductor.org/packages/Omixer/), has been made available [@sinke2021].
-    It tests if the random assignments are homogeneous transforming all variables to numeric and using the Kendall's correlation if there are more than 5 samples, otherwise using Pearson's chi-squared test.
+-   Recently, *Omixer*, a new package, has recently been made available at [Bioconductor](https://bioconductor.org/packages/Omixer/) [@sinke2021].
+    It tests whether the random assignments are homogeneous by transforming all variables to numeric values and using the Kendall's correlation when there are more than 5 samples; otherwise , it utilizes the Pearson's chi-squared test.
 
--   Finally the package *experDesign*, on [CRAN](https://cran.r-project.org/package=experDesign), which provides groups with characteristics similar to the whole sample by comparing the random samples with the whole dataset for several statistics.
+-   Finally there is the package *experDesign*, at [CRAN](https://cran.r-project.org/package=experDesign), which provides groups with characteristics similar to the entire sample by comparing the random samples with the whole dataset across several statistics.
 
@@
 
-The package ***experDesign*** provides the function `design` to distribute the samples on multiple batches so that each variable is homogeneous in each batch.
-It is similar to the *anticluster* method for maximum variance.
-If the experiment is carried out on a specific spatial distribution the `spatial` function distributes the samples homogeneously also by position similar to *Omixer*.
+The package ***experDesign*** provides the functional `design` to distribute the samples into multiple batches such that each variable is homogeneous within each batch.
+It is similar to the *anticluster* method used for maximum variance.
+If the experiment is carried out with a specific spatial distribution, the `spatial` function also distributes the samples homogeneously by position in a manner similar to *Omixer*.
 
-In addition to distribute the samples on batches ***experDesign*** provides with tools to add technical replicates.
-When the design is already done then the replicates are called technical replicates [@blainey2014].
-Technical replicates are samples that are measured multiple times and allow to reduce the uncertainty of the measurement.
-If the technical replicates are distributed on several batches they allow to measure the batch effect and thus minimize it.
-To select the technical replicates needed and choose from which samples they are done the function `extreme_cases` is provided.
-For easier usage the `replicates` designs an experiment with the number of replicates per batch desired.
+In addition to distributing the samples into batches, ***experDesign*** provides tools to add technical replicates.
+When the design is complete, the replicates are referred to as technical replicates [@blainey2014].Technical replicates are samples that are measured multiple times, which reduces the uncertainty of the measurement.
+If the technical replicates are distributed across several batches, batch effects can be measured and thus minimized.
+To select the technical replicates needed and in order to choose from which samples they are calculated, the function `extreme_cases` is provided.
+For easier usage, the `replicates` function designs an experiment with the number of replicates per batch desired.
 
-***experDesign*** also provides several small utilities to make it easier design the experiment in batches.
-For instance, it helps to calculate the samples to distribute on the minimum number of batches required with `sizes_batches`.
+***experDesign*** also provides several small utilities to make it easier to design the experiment in batches.
+For instance, a function called `sizes_batches` helps calculate the number of samples in order to distribute them across the number of batches required.
 

### 2021-07-12 paper.Rmd
Add acknowledgments
Some changes from reviewers

@@
 ---
-title: 'experDesign: helping performing experiments on batches'
+title: 'experDesign: stratifyuing samples into batches with minimal bias'
 output:
+  html_document:
+    df_print: paged
   md_document:
     preserve_yaml: yes
-  html_document:
-    df_print: paged
   word_document:
@@
 
+```{r setup}
+knitr::opts_chunk$set(collapse = TRUE)
+```
+
 # Summary
@@
 Failure to take such variability into account can render an experiment inconclusive.
-*experDesign* provides tools to minimize the risk of inconclusive results by specifically assigning samples to particular batches.
+*experDesign* provides tools to minimize the risk of inconclusive results by assigning samples to batches to minimize potential batch effects.
 
@@
 Sometimes, however, this is not possible, and then one needs to apply certain techniques to control for such variations.
-There are three techniques used to reduce unwanted variation: blocking, randomization and replication [@klaus2015].
+There are three techniques used to decrease the uncertainty about the unwanted variation: blocking, randomization and replication [@klaus2015].
 
-**Blocking** group samples that are equal, according to one or more variables, allows one to estimate the variation in the measurements of the factors from these variables.
-**Randomization** is a method to minimize the variation in the measurements by mixing the potential confounding variables.
+**Blocking** groups samples that are equal according to one or more variables allowing one to estimate the differences between each batch by comparing measurements within blocks.
+**Randomization** by mixing the potential confounding variables minimizes the variation in the measurements .
 **Replication** increases the number of samples used in an experiment to better estimate the variation of the experiment.
-In some settings (e.g., clinical, agriculture, etc.), several of these techniques can be applied together to ensure the robustness of the study.
+In some settings these techniques can be applied together to ensure the robustness of the study.
 
-Between the designing of an experiment and the measurement of the samples, many things can happen.
-If some samples are lost, are contaminated, or do not pass quality control for measurement purposes, the unwanted variation of the experiment will increase.
-Even if this doesn't happen, experiments will occasionally need to be carried out in batches due to technical reasons; for example, the machine cannot measure more than one samples at a time, or because of other practical considerations; for instance, it may not be possible to obtain additional measurements in the field during the allotted time.
+Between the designing of an experiment and the measurement of the samples, some samples might be lost, be contaminated, or degradation below quality control threshold can happen.
+Even if this does not happen, experiments will occasionally need to be carried out in batches.
+Due to technical, for example, the machine cannot measure more than one sample at a time; and practical reasons, for instance, it may not be possible to obtain additional measurements in the field during the allotted time.
 
+This divergence from the original design might cause batch effects perturbing the analysis.
 There are several techniques to identify and assess batch effects when analyzing an already measured experiment [@leek2010].
-Experiments that run over long periods of time or that are run across different laboratories are highly susceptible to batch effects.
-This is also true of even smaller single-laboratory studies, if they span several days or include personnel changes.
-However, if the source of variations are not taken into account before measuring is carried out, a batch effect can be introduced that later is almost impossible to remove.
-Thus, it would be better to avoid such batch effects before carrying out an experiment.
-Taking into account the process from the design phase to the final measurement can help prevent batch effects.
+However, if the source of variations is not taken into account before measurement is completed, a batch effect can be introduced that confounds the analysis.
+Thus, it would be better to avoid such batch effects before executing an experiment.
 
-Experiments should be designed to distribute batches and other potential sources of experimental variation across groups.
-To prevent the occurrence of batch effects after the initial design of the experiment, there are two options: randomization and replication.
-These techniques can not only prevent batch effects, but can also facilitate comparisons between or within the desired variable(s).
-Both can be utilized, but in any case one must take blocking into account from the beginning or the problem might be further exacerbated.
-
-Randomization, if done correctly, can helps reducing variations across groups, while replication can help one be more confident of the experimental results.
-For instance, if one designs an experiment with cases and controls and the latter are measured in one batch and the controls in a different batch, then any difference between them can be entirely confounded by a malfunction of the measuring machine in either one of the batches [@chen2011].
-However, by examining how the variables are distributed across each batch sample, proper randomized can be ensured, thus minimizing batch effects.
-This is known as randomized block experimental design or stratified random sampling experimental design.
-
+To prevent the batch effect to confound your analysis after the initial design of the experiment, there are two options: randomization and replication.
+Randomization, can help reduce variations across groups, while replication helps estimate the variation of the measurements or samples, thus increasing the precision of the estimates of the true value on the analysis.
 Replications consist of increasing the number of measurements with similar attributes.
-Usually at least three samples are included for each condition of interest in order to estimate the variations of these attributes.
 When there is one extraction and then a sample is measured multiple times, this is referred to as a technical replicate.
@@
 
-Replications consist on increasing the number of measurements with similar attributes.
-Usually at least three samples are included for each condition of interest to be able to estimate the variation of the attributes.
-If there is one extraction and then a sample is measured multiple times it is called a technical replicate.
-Technical replicates help estimate the variation of the measurement method and thus of the possible batch effect.
+Randomization and replicates can be utilized to prevent batch effects to confound the analysis.
+However, by examining how the variables are distributed across each batch sample, proper randomization can be ensured, thus minimizing batch effects.
+This is known as randomized block experimental design or stratified random sampling experimental design.
 
@@
 There are some tools to prevent batch effects on the R language in multiple fields and areas, and particularly for biological research [@rcoreteam2014].
-They carry some caveats limiting their application in some cases, and to our knowledge no comparisons of these various methods has been conducted.
 Here we briefly describe the currently available packages:
@@
 -   Recently, *Omixer*, a new package, has recently been made available at [Bioconductor](https://bioconductor.org/packages/Omixer/) [@sinke2021].
-    It tests whether the random assignments are homogeneous by transforming all variables to numeric values and using the Kendall's correlation when there are more than 5 samples; otherwise , it utilizes the Pearson's chi-squared test.
+    It tests whether the random assignments are homogeneous by transforming all variables to numeric values and using the Kendall's correlation when there are more than 5 samples; otherwise, it utilizes the Pearson's chi-squared test.
 
--   Finally there is the package *experDesign*, at [CRAN](https://cran.r-project.org/package=experDesign), which provides groups with characteristics similar to the entire sample by comparing the random samples with the whole dataset across several statistics.
+# Statement of need
+
+Current solutions to stratify samples to reduce and control batch effect do not work for all cases and usual needs of researchers.
+They are either specialized to specific type of data or omit some conditions that are often meet or only work on a specific subset of conditions.
+The new package ***experDesign*** works on all kind of data types and doesn't require a spatial distribution making it suitable for all kind of experiments.
+This package is intended for people looking for a quick easy to interpret solution providing reasonable suggestions about how to distribute the samples on their analysis.
+
+For completeness a description and comparison of the usage of the different software currently available on CRAN and Bioconductor is presented below.
+First some real data from a survey:
+
+```{r example-data}
+data(survey, package = "MASS")
+survey$ID <- seq_len(nrow(survey))
+VoI <- c("Sex", "Smoke", "Age")
+n_batch <- 3
+size_subset <- 96
+iterations <- 1000
+```
+
+This data set has three variables of interest; Sex, Smoke and Age are a mix of categorical and numeric variables.
+
+## OSAT
+
+OSAT provides some template for some setups and works for numeric and categorical variables:
+
+```{r osat}
+library("ggplot2")
+library("OSAT")
+library("patchwork")
+gs <- setup.sample(survey, optimal = VoI)
+gc <- setup.container(plate = IlluminaBeadChip96Plate, n = n_batch, batch = 'plates')
+gSetup <- create.optimized.setup(sample = gs, container = gc, nSim = iterations)
+osat_report <- get.experiment.setup(gSetup)
+nrow(osat_report)
+osat_age <- ggplot(osat_report) +
+  geom_histogram(aes(Age, fill = as.character(plates))) +
+  facet_wrap(~plates) +
+  labs(fill = "Batch")
+osat_sex <- ggplot(osat_report) +
+  geom_bar(aes(Sex, fill = as.character(plates))) +
+  facet_wrap(~plates) +
+  labs(fill = "Batch")
+osat_smoke <- ggplot(osat_report) +
+  geom_bar(aes(Smoke, fill = as.character(plates))) +
+  facet_wrap(~plates) +
+  labs(fill = "Batch")
+(osat_smoke/(osat_age + osat_sex + plot_layout(guides = 'collect'))) +
+  plot_annotation(title = "OSAT")
+```
+
+OSAT returns one row less that the input provided because it deletes one row because it has an NA value on the Sex variable.
+
+## anticlust
+
+anticlust doesn't handle all the variables, it only accepts numeric variables:
+
+```{r anticlust, error=TRUE}
+library("anticlust")
+anticlust_index <- anticlustering(
+  survey[, VoI],
+  K = n_batch,
+  objective = "variance",
+  method = "exchange",
+  repetitions = iterations
+)
+anticlust_index <- anticlustering(
+  survey[, "Age"],
+  K = n_batch,
+  objective = "variance",
+  method = "exchange",
+  repetitions = iterations
+)
+report_anticlust <- data.frame(Age = survey[, "Age"], batch = as.character(anticlust_index))
+library("ggplot2")
+ggplot(report_anticlust) + 
+  geom_histogram(aes(Age, fill = batch)) +
+  facet_wrap(~batch) +
+  labs(fill = "Batch", title = "anticlust")
+```
+
+## Omixer
+
+There is a bug on Omixer that prevents it to work unless specific conditions are meet that prevents it comparing with the other tools using the same settings:
+
+```{r Omixer, error=TRUE}
+library("Omixer")
+data(survey, package = "MASS")
+survey$ID <- seq_len(nrow(survey))
+Omixer_index <- omixerRand(survey, sampleId = "ID", 
+                           block = "block", iterNum = 10, wells = size_subset, 
+                           plateNum = n_batch, randVars = VoI, techVars = NULL)
+Omixer_index <- omixerRand(survey, sampleId = "ID", 
+                           block = "block", iterNum = 10, wells = size_subset, mask = rep(1, nrow(survey)),
+                           plateNum = n_batch, randVars = VoI, techVars = NULL)
+Omixer_index <- omixerRand(survey, sampleId = "ID", 
+                           block = "block", iterNum = 10, wells = size_subset, mask = rep(0, nrow(survey)),
+                           plateNum = n_batch, randVars = VoI, techVars = NULL)
+Omixer_index <- omixerRand(survey, sampleId = "ID", 
+                           block = "block", iterNum = 10, wells = size_subset, mask = rep(1, 288),
+                           plateNum = n_batch, randVars = VoI, techVars = NULL)
+Omixer_index <- omixerRand(survey, sampleId = "ID", 
+                           block = "block", iterNum = 10, wells = size_subset, mask = rep(0, 288),
+                           plateNum = n_batch, randVars = VoI, techVars = NULL)
+
+```
+
+## experDesign
+
+***experDesign***, the package described on this paper, first evaluates the distribution of samples in batches and then on a spatial grid.
+
+```{r experDesign}
+library("experDesign")
+experDesign_index <- design(survey[, VoI], size_subset = size_subset,
+                  iterations = iterations)
+experDesign_index_spatial <- spatial(index = experDesign_index, pheno = survey[, VoI], 
+                                     iterations = iterations, rows = LETTERS[1:8], columns = 1:12)
+experDesign_report <- inspect(experDesign_index, survey[, VoI])
+nrow(experDesign_report)
+experDesign_age <- ggplot(experDesign_report) + 
+  geom_histogram(aes(Age, fill = batch)) +
+  facet_wrap(~batch) +
+  labs(fill = "Batch")
+experDesign_sex <- ggplot(experDesign_report) +
+  geom_bar(aes(Sex, fill = batch)) +
+  facet_wrap(~batch) +
+  labs(fill = "Batch")
+experDesign_smoke <- ggplot(experDesign_report) +
+  geom_bar(aes(Smoke, fill = batch)) +
+  facet_wrap(~batch) +
+  labs(fill = "Batch")
+(experDesign_smoke/(experDesign_age + experDesign_sex + plot_layout(guides = 'collect'))) + plot_annotation(title = "experDesign")
+```
 
@@
 The package ***experDesign*** provides the functional `design` to distribute the samples into multiple batches such that each variable is homogeneous within each batch.
-It is similar to the *anticluster* method used for maximum variance.
-If the experiment is carried out with a specific spatial distribution, the `spatial` function also distributes the samples homogeneously by position in a manner similar to *Omixer*.
+Each batch is set to have some centrality and dispersion statistics to match as closely as possible the original input design data.
+It does by comparing the mean, the standard deviation, the median absolute deviation, variables with no value number, the entropy and the independence of the categorical variables.
+The distribution of samples that more closely match the original distribution is returned to the user.
+
+This approach is similar to the *anticlust* method used for maximum variance, where the objective function is the sum of the squared errors between cluster centers and individual data points.
+But instead of comparing to the clusters center it is compared to the whole data.
+If the experiment is carried out with a specific spatial distribution, the `spatial` function also distributes the samples homogeneously by position.
 
 In addition to distributing the samples into batches, ***experDesign*** provides tools to add technical replicates.
-When the design is complete, the replicates are referred to as technical replicates [@blainey2014].Technical replicates are samples that are measured multiple times, which reduces the uncertainty of the measurement.
-If the technical replicates are distributed across several batches, batch effects can be measured and thus minimized.
+When the design is complete, the replicates are referred to as technical replicates [@blainey2014].
 To select the technical replicates needed and in order to choose from which samples they are calculated, the function `extreme_cases` is provided.
 For easier usage, the `replicates` function designs an experiment with the number of replicates per batch desired.
+
+```{r controls}
+extreme_cases(survey[, VoI], size = 5)
+replicates_index <- replicates(survey[, VoI], size_subset = size_subset, controls = 5, iterations = iterations)
+spatial_replicate_index <- spatial(replicates_index, survey[, VoI])
+inspect(spatial_replicate_index, survey[, VoI])
+```
 
@@
 
+```{r show-batches}
+optimum_batches(size_data = 250, size_subset = 96)
+sizes_batches(size_data = 250, size_subset = 96, batches = 3)
+```
+
+# Acknowledgments
+
+We are grateful to Joe Moore for English-language assistance.
+
 # References

### 2021-07-15 paper.Rmd
Update text

@@
 output:
+  word_document:
+    keep_md: yes
+  md_document:
+    preserve_yaml: yes
+  pdf_document: default
   html_document:
     df_print: paged
-  md_document:
-    preserve_yaml: yes
-  word_document:
-    keep_md: yes
-date: "23 April 2021"
+date: "15th July 2021"
 affiliations:
@@
 
-```{r setup}
-knitr::opts_chunk$set(collapse = TRUE)
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(collapse = TRUE, warning = TRUE, error = TRUE)
 ```
@@
 
-Randomization and replicates can be utilized to prevent batch effects to confound the analysis.
+Randomization and replicates can be used to prevent batch effects to confound the analysis.
 However, by examining how the variables are distributed across each batch sample, proper randomization can be ensured, thus minimizing batch effects.
@@
 
-There are some tools to prevent batch effects on the R language in multiple fields and areas, and particularly for biological research [@rcoreteam2014].
+There are some tools to minimize batch effects on the R language in multiple fields and areas, and particularly for biological research [@rcoreteam2014].
 Here we briefly describe the currently available packages:
@@
 library("Omixer")
-data(survey, package = "MASS")
 survey$ID <- seq_len(nrow(survey))
@@
 
-## experDesign
+# Description
 
-***experDesign***, the package described on this paper, first evaluates the distribution of samples in batches and then on a spatial grid.
+The package ***experDesign*** provides the functional `design` to distribute the samples into multiple batches such that each variable is homogeneous within each batch.
+Each batch is set to have some centrality and dispersion statistics to match as closely as possible the original input design data.
+It does by comparing the mean, the standard deviation, the median absolute deviation, variables with no value number, the entropy and the independence of the categorical variables.
+On each iteration if the distribution of samples for each batch has less differences with the original distribution than the last previous stored sample distribution then it replace it as the best sample distribution.
+At the end of the iterations the best sample distribution is returned to the user.
 
@@
 
-# Description
+One can check the statistics of the index for multiple statistics:
 
-The package ***experDesign*** provides the functional `design` to distribute the samples into multiple batches such that each variable is homogeneous within each batch.
-Each batch is set to have some centrality and dispersion statistics to match as closely as possible the original input design data.
-It does by comparing the mean, the standard deviation, the median absolute deviation, variables with no value number, the entropy and the independence of the categorical variables.
-The distribution of samples that more closely match the original distribution is returned to the user.
+```{r evaluate}
+evaluate_na(experDesign_index, survey[, VoI])
+evaluate_entropy(experDesign_index, survey[, VoI])
+evaluate_mad(experDesign_index, survey[, VoI])
+evaluate_sd(experDesign_index, survey[, VoI])
+evaluate_mean(experDesign_index, survey[, VoI])
+# All together for each batch
+ei <- evaluate_index(experDesign_index, survey[, VoI])
+ei[, , "SubSet1"]
+evaluate_na(experDesign_index_spatial, survey[, VoI])
+evaluate_entropy(experDesign_index_spatial, survey[, VoI])
+evaluate_mad(experDesign_index_spatial, survey[, VoI])
+evaluate_sd(experDesign_index_spatial, survey[, VoI])
+evaluate_mean(experDesign_index_spatial, survey[, VoI])
+# All together for each batch
+ei <- evaluate_index(experDesign_index_spatial, survey[, VoI])
+ei[, , "G1"]
+```
 
-This approach is similar to the *anticlust* method used for maximum variance, where the objective function is the sum of the squared errors between cluster centers and individual data points.
+And compare it with the original distribution:
+
+```{r original}
+evaluate_orig(survey[, VoI])
+```
+
+In ***experDesign*** the difference is used to compare with the original distribution, this approach is similar to the *anticlust* method used for maximum variance, but instead of the difference with the original distribution the objective function is the sum of the squared errors between cluster centers and individual data points.
 But instead of comparing to the clusters center it is compared to the whole data.
 If the experiment is carried out with a specific spatial distribution, the `spatial` function also distributes the samples homogeneously by position.
+See below for an example.
 
@@
 extreme_cases(survey[, VoI], size = 5)
-replicates_index <- replicates(survey[, VoI], size_subset = size_subset, controls = 5, iterations = iterations)
-spatial_replicate_index <- spatial(replicates_index, survey[, VoI])
-inspect(spatial_replicate_index, survey[, VoI])
+replicates_index <- replicates(pheno = survey[, VoI], 
+                               size_subset = size_subset, controls = 5, 
+                               iterations = iterations)
+spatial_replicate_index <- spatial(index = replicates_index, pheno = survey[, VoI], 
+                                   rows = LETTERS[1:8], columns = 1:12)
+survey$ID <- seq_len(nrow(survey))
+report_replicates <- inspect(i = replicates_index, 
+                             pheno = survey[, c(VoI, "ID")])
+report_replicates_position <- inspect(i = spatial_replicate_index, 
+                             pheno = report_replicates, 
+                             index_name = "position")
+# Batch and position where the samples should be:
+head(report_replicates_position)
+# Technical replicates:
+head(report_replicates_position[duplicated(report_replicates_position$ID), ])
 ```
@@
 ***experDesign*** also provides several small utilities to make it easier to design the experiment in batches.
-For instance, a function called `sizes_batches` helps calculate the number of samples in order to distribute them across the number of batches required.
+For instance, a function called `sizes_batches` helps calculate the number of samples in order to distribute them across the number of batches required and another one calculates the minimal amount of batches required:
 
@@
 
+In conclusion the ***experDesign*** can be a first fast step to prepare for a batch experiment.
+It can use as many numeric and categorical variables as needed to stratify the experiment design on batches including spatial distributions.
+
 # Acknowledgments

### 2021-07-20 paper.Rmd
Update paper and add figures

@@
 ---
-title: 'experDesign: stratifyuing samples into batches with minimal bias'
+title: 'experDesign: stratifying samples into batches with minimal bias'
 output:
@@
 Sometimes, however, this is not possible, and then one needs to apply certain techniques to control for such variations.
-There are three techniques used to decrease the uncertainty about the unwanted variation: blocking, randomization and replication [@klaus2015].
+There are three techniques used to decrease the uncertainty of the unwanted variation: blocking, randomization and replication [@klaus2015].
 
-**Blocking** groups samples that are equal according to one or more variables allowing one to estimate the differences between each batch by comparing measurements within blocks.
-**Randomization** by mixing the potential confounding variables minimizes the variation in the measurements .
+**Blocking** groups samples that are equal according to one or more variables, allowing one to estimate the differences between each batch by comparing measurements within blocks.
+**Randomization** minimizes the variation in the measurements by mixing the potential confounding variables.
 **Replication** increases the number of samples used in an experiment to better estimate the variation of the experiment.
@@
 
-Between the designing of an experiment and the measurement of the samples, some samples might be lost, be contaminated, or degradation below quality control threshold can happen.
+Between the designing of an experiment and the measurement of the samples, some samples might be lost, be contaminated, or degradation below quality threshold can happen.
 Even if this does not happen, experiments will occasionally need to be carried out in batches.
-Due to technical, for example, the machine cannot measure more than one sample at a time; and practical reasons, for instance, it may not be possible to obtain additional measurements in the field during the allotted time.
+Batches might be due to technical reasons, for example, the machine cannot measure more than some amount of samples at a time; and practical reasons, for instance, it may not be possible to obtain additional measurements in the field during the allotted time.
 
@@
 
-To prevent the batch effect to confound your analysis after the initial design of the experiment, there are two options: randomization and replication.
+To prevent the batch effect to confound an analysis after the initial design of the experiment, there are two options: randomization and replication.
 Randomization, can help reduce variations across groups, while replication helps estimate the variation of the measurements or samples, thus increasing the precision of the estimates of the true value on the analysis.
 Replications consist of increasing the number of measurements with similar attributes.
-When there is one extraction and then a sample is measured multiple times, this is referred to as a technical replicate.
-Technical replicates help estimate the variation of the measurement method, and thus the possible batch effect.
+When a sample is measured multiple times, this is referred to as a technical replicate.
+Technical replicates help estimate the variation of the measurement method, and thus the possible batch effect [@blainey2014].
 
 Randomization and replicates can be used to prevent batch effects to confound the analysis.
-However, by examining how the variables are distributed across each batch sample, proper randomization can be ensured, thus minimizing batch effects.
+However, by examining how the variables are distributed across each batch, proper randomization can be ensured, thus minimizing batch effects.
 This is known as randomized block experimental design or stratified random sampling experimental design.
@@
 
--   *minDiff*, at [github](https://github.com/m-Py/minDiff), and its successor *anticlust*, at [CRAN](https://cran.r-project.org/package=anticlust), divide the samples into similar groups, ensuring similarity by enforcing heterogeneity within groups [@papenberg2020].
+-   *minDiff*, at [github](https://github.com/m-Py/minDiff), and its successor *anticlust*, at [CRAN](https://cran.r-project.org/package=anticlust), divides the samples into similar groups, ensuring similarity by enforcing heterogeneity within groups [@papenberg2020].
     Conceptually it is similar to the clustering methods k-means.
@@
 
-There is a bug on Omixer that prevents it to work unless specific conditions are meet that prevents it comparing with the other tools using the same settings:
+There is a bug on Omixer that prevents it to work unless specific conditions are meet.
+That prevents it comparing with the other tools using the same settings:
 
@@
 
-In ***experDesign*** the difference is used to compare with the original distribution, this approach is similar to the *anticlust* method used for maximum variance, but instead of the difference with the original distribution the objective function is the sum of the squared errors between cluster centers and individual data points.
-But instead of comparing to the clusters center it is compared to the whole data.
+In ***experDesign*** the difference is used to compare with the original distribution, this approach is similar to the *anticlust* method used for maximum variance, but instead of the difference with the original distribution the objective function is the sum of the squared errors between cluster centers and individual data points and instead of comparing to the clusters center it is compared to the whole data.
 If the experiment is carried out with a specific spatial distribution, the `spatial` function also distributes the samples homogeneously by position.
@@
 In addition to distributing the samples into batches, ***experDesign*** provides tools to add technical replicates.
-When the design is complete, the replicates are referred to as technical replicates [@blainey2014].
-To select the technical replicates needed and in order to choose from which samples they are calculated, the function `extreme_cases` is provided.
+In order to choose them from the available samples, the function `extreme_cases` is provided.
 For easier usage, the `replicates` function designs an experiment with the number of replicates per batch desired.

### 2021-07-26 paper.Rmd
Comments Anna

@@
     df_print: paged
-date: "15th July 2021"
+date: "26st July 2021"
 affiliations:
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(collapse = TRUE, warning = TRUE, error = TRUE)
+knitr::opts_chunk$set(collapse = TRUE, warning = TRUE, error = TRUE, fig.width = 9)
 ```
@@
 Typically, one can control the environment in which the study or experiment is being conducted.
-Sometimes, however, this is not possible, and then one needs to apply certain techniques to control for such variations.
+Sometimes, however, this is not possible.
+In this event, techniques to control such variations need to be applied.
 There are three techniques used to decrease the uncertainty of the unwanted variation: blocking, randomization and replication [@klaus2015].
 
-**Blocking** groups samples that are equal according to one or more variables, allowing one to estimate the differences between each batch by comparing measurements within blocks.
-**Randomization** minimizes the variation in the measurements by mixing the potential confounding variables.
+**Blocking** groups samples that are equal according to one or more variables, allowing the estimation of the differences between each batch by comparing measurements within blocks.
+**Randomization** minimizes the variation in the measurements by randomly mixing the potential confounding variables.
 **Replication** increases the number of samples used in an experiment to better estimate the variation of the experiment.
@@
 
-Between the designing of an experiment and the measurement of the samples, some samples might be lost, be contaminated, or degradation below quality threshold can happen.
+Between the designing of an experiment and the measurement of the samples, some samples might be lost, contaminated, or degradated below quality threshold.
 Even if this does not happen, experiments will occasionally need to be carried out in batches.
-Batches might be due to technical reasons, for example, the machine cannot measure more than some amount of samples at a time; and practical reasons, for instance, it may not be possible to obtain additional measurements in the field during the allotted time.
+Batches might be due to technical reasons, for example, the machine cannot measure more than a number of samples at the same time.
+They can be also due to practical reasons, for instance, it may not be possible to obtain additional measurements in the field during the allotted time.
 
@@
 
--   *minDiff*, at [github](https://github.com/m-Py/minDiff), and its successor *anticlust*, at [CRAN](https://cran.r-project.org/package=anticlust), divides the samples into similar groups, ensuring similarity by enforcing heterogeneity within groups [@papenberg2020].
+-   *anticlust*, at [CRAN](https://cran.r-project.org/package=anticlust), divides the samples into similar groups, ensuring similarity by enforcing heterogeneity within groups [@papenberg2020].
     Conceptually it is similar to the clustering methods k-means.
 
--   Recently, *Omixer*, a new package, has recently been made available at [Bioconductor](https://bioconductor.org/packages/Omixer/) [@sinke2021].
-    It tests whether the random assignments are homogeneous by transforming all variables to numeric values and using the Kendall's correlation when there are more than 5 samples; otherwise, it utilizes the Pearson's chi-squared test.
+-   Recently, *Omixer*, a new package, has been made available at [Bioconductor](https://bioconductor.org/packages/Omixer/) [@sinke2021].
+    It tests whether the random assignments made by it are homogeneous by transforming all variables to numeric values and using the Kendall's correlation when there are more than 5 samples; otherwise, it utilizes the Pearson's chi-squared test.
 
@@
 Current solutions to stratify samples to reduce and control batch effect do not work for all cases and usual needs of researchers.
-They are either specialized to specific type of data or omit some conditions that are often meet or only work on a specific subset of conditions.
+They are either specialized to specific type of data, omit some conditions that are often meet or only work on a specific subset of conditions.
 The new package ***experDesign*** works on all kind of data types and doesn't require a spatial distribution making it suitable for all kind of experiments.
-This package is intended for people looking for a quick easy to interpret solution providing reasonable suggestions about how to distribute the samples on their analysis.
+This package is intended for people looking for a quick and easy to interpret solution providing reasonable suggestions about how to distribute the samples on their analysis.
 
@@
 
-```{r osat}
+```{r osat, fig.cap="Samples distributions with OSAT."}
 library("ggplot2")
+theme_plots <- theme_minimal() + theme(text = element_text(size = 10))
+theme_set(theme_plots)
 library("OSAT")
@@
 
-```{r anticlust, error=TRUE}
+```{r anticlust, error=TRUE, fig.cap="Samples distributions with anticlust."}
 library("anticlust")
@@
 
-The package ***experDesign*** provides the functional `design` to distribute the samples into multiple batches such that each variable is homogeneous within each batch.
+The package ***experDesign*** provides the function `design` to arrange the samples into multiple batches such that the variable's distribution is homogeneous within each batch.
 Each batch is set to have some centrality and dispersion statistics to match as closely as possible the original input design data.
-It does by comparing the mean, the standard deviation, the median absolute deviation, variables with no value number, the entropy and the independence of the categorical variables.
-On each iteration if the distribution of samples for each batch has less differences with the original distribution than the last previous stored sample distribution then it replace it as the best sample distribution.
+The statistics used are the mean, the standard deviation, the median absolute deviation, variables with no value number, the entropy and the independence of the categorical variables.
+On each iteration if the random distribution of samples' statitics for each batch has less differences with the original distribution than the last stored sample distribution then it replaces it as the best sample distribution.
 At the end of the iterations the best sample distribution is returned to the user.
 
-```{r experDesign}
+Users can follow the following flowchart to decide what function(s) they need to use:
+
+![Flow chart to decide which functions are needed](flow_chart.png){width="353"}
+
+If users want a design without replicates but batches have some spatial distribution, we need to use `design` to allocate the samples on each batch and then the `spatial` function to randomly distribute the samples homogeneously by position within each batch.
+The samples are distributed on each position as with the `design` function, only returning the distribution with less differences between the statistics previously mentioned with the original distribution to the user of all the iterations tried.
+See below for an example:
+
+```{r experDesign, fig.cap="Samples distributions with experDesign."}
 library("experDesign")
 experDesign_index <- design(survey[, VoI], size_subset = size_subset,
-                  iterations = iterations)
+                  iterations = iterations, name = "")
 experDesign_index_spatial <- spatial(index = experDesign_index, pheno = survey[, VoI], 
@@
 
-One can check the statistics of the index for multiple statistics:
+The statistics of the index can be checked for multiple statistics, as shown in the following code:
 
@@
 
-And compare it with the original distribution:
+We can also compare our results with the original distribution:
 
@@
 
-In ***experDesign*** the difference is used to compare with the original distribution, this approach is similar to the *anticlust* method used for maximum variance, but instead of the difference with the original distribution the objective function is the sum of the squared errors between cluster centers and individual data points and instead of comparing to the clusters center it is compared to the whole data.
-If the experiment is carried out with a specific spatial distribution, the `spatial` function also distributes the samples homogeneously by position.
-See below for an example.
-
 In addition to distributing the samples into batches, ***experDesign*** provides tools to add technical replicates.
@@
 For easier usage, the `replicates` function designs an experiment with the number of replicates per batch desired.
+The rest of the workflow is similar to the above example:
 
@@
 ***experDesign*** also provides several small utilities to make it easier to design the experiment in batches.
-For instance, a function called `sizes_batches` helps calculate the number of samples in order to distribute them across the number of batches required and another one calculates the minimal amount of batches required:
+For instance, a function called `sizes_batches` helps calculate the number of samples in order to distribute them across the batches required.
+Furthermore, `optimum_batches` calculates the minimal amount of batches required:
 
@@
 
-In conclusion the ***experDesign*** can be a first fast step to prepare for a batch experiment.
+In conclusion the ***experDesign*** can be a first and fast step to prepare for a batched experiment.
 It can use as many numeric and categorical variables as needed to stratify the experiment design on batches including spatial distributions.

### 2021-07-26 paper.Rmd
Changes Azucena

@@
 
-To properly design an experiment, the source of the variation between samples must be identified.
+To design an experiment that can support conclusive results upon analysis, the source of the variation between samples must be identified.
 Typically, one can control the environment in which the study or experiment is being conducted.
@@
 
-**Blocking** groups samples that are equal according to one or more variables, allowing the estimation of the differences between each batch by comparing measurements within blocks.
+**Blocking** is a method that groups samples that are equal according to one or more variables, allowing the estimation of the differences between each batch by comparing measurements within blocks.
 **Randomization** minimizes the variation in the measurements by randomly mixing the potential confounding variables.
@@
 Between the designing of an experiment and the measurement of the samples, some samples might be lost, contaminated, or degradated below quality threshold.
-Even if this does not happen, experiments will occasionally need to be carried out in batches.
-Batches might be due to technical reasons, for example, the machine cannot measure more than a number of samples at the same time.
+In addition, experiments will occasionally need to be carried out in batches.
+Batches might be necessary for technical reasons, for example, the device cannot measure more than a number of samples at the same time.
 They can be also due to practical reasons, for instance, it may not be possible to obtain additional measurements in the field during the allotted time.
@@
 There are several techniques to identify and assess batch effects when analyzing an already measured experiment [@leek2010].
-However, if the source of variations is not taken into account before measurement is completed, a batch effect can be introduced that confounds the analysis.
-Thus, it would be better to avoid such batch effects before executing an experiment.
+It would be better to avoid such batch effects before executing an experiment.
+By taking into account the de differences between the original design and the state before measurement is done, confounding effects could be minimized.
 
-To prevent the batch effect to confound an analysis after the initial design of the experiment, there are two options: randomization and replication.
-Randomization, can help reduce variations across groups, while replication helps estimate the variation of the measurements or samples, thus increasing the precision of the estimates of the true value on the analysis.
+To prevent the batch effect fom confounding the analysis after the initial design of the experiment, there are two options: randomization and replication.
+Randomization, consists on shuffling the samples to mix samples with different attributes, can help reduce variations across groups, while replication helps estimate the variation of the measurements or samples, thus increasing the precision of the estimates of the true value on the analysis.
 Replications consist of increasing the number of measurements with similar attributes.
@@
 Randomization and replicates can be used to prevent batch effects to confound the analysis.
-However, by examining how the variables are distributed across each batch, proper randomization can be ensured, thus minimizing batch effects.
+By examining how the variables are distributed across each batch, proper randomization can be ensured, thus minimizing batch effects.
 This is known as randomized block experimental design or stratified random sampling experimental design.
@@
 
-There are some tools to minimize batch effects on the R language in multiple fields and areas, and particularly for biological research [@rcoreteam2014].
+There are some tools to minimize batch effects on the R language in multiple fields, and particularly for biological research [@rcoreteam2014].
 Here we briefly describe the currently available packages:
@@
 
-Current solutions to stratify samples to reduce and control batch effect do not work for all cases and usual needs of researchers.
-They are either specialized to specific type of data, omit some conditions that are often meet or only work on a specific subset of conditions.
-The new package ***experDesign*** works on all kind of data types and doesn't require a spatial distribution making it suitable for all kind of experiments.
-This package is intended for people looking for a quick and easy to interpret solution providing reasonable suggestions about how to distribute the samples on their analysis.
+Current solutions to stratify samples to reduce and control batch effect do not work for all cases.
+They are either specialized to particular type of data, they omit some conditions that are often met or they only work on a specific subset of conditions.
+The new package ***experDesign*** works on all kind of data types and does not require a spatial distribution making it suitable for all kind of experiments.
+This package is intended for people looking for a quick and easy solution providing reasonable suggestions on how to distribute the samples for their analysis.
 
 For completeness a description and comparison of the usage of the different software currently available on CRAN and Bioconductor is presented below.
-First some real data from a survey:
+First, we start with some real data from a survey:
 
@@
 
-OSAT provides some template for some setups and works for numeric and categorical variables:
+OSAT provides some template for plates that hold 2, 4, 8 Illumina BeadChip chips and have 24, 48, 96 wells and works for numeric and categorical variables:
 
@@
 
-OSAT returns one row less that the input provided because it deletes one row because it has an NA value on the Sex variable.
+OSAT returns one row less that the input provided.
+This row is deleted because it has an NA value on the Sex variable.
 
@@
 
-anticlust doesn't handle all the variables, it only accepts numeric variables:
+anticlust does not handle all the variables, it only accepts numeric variables:
 
@@
 The statistics used are the mean, the standard deviation, the median absolute deviation, variables with no value number, the entropy and the independence of the categorical variables.
-On each iteration if the random distribution of samples' statitics for each batch has less differences with the original distribution than the last stored sample distribution then it replaces it as the best sample distribution.
+On each iteration if the random distribution of samples' statitics for each batch has fewer differences with the original distribution than the last stored sample distribution then it replaces it as the best sample distribution.
 At the end of the iterations the best sample distribution is returned to the user.
@@
 If users want a design without replicates but batches have some spatial distribution, we need to use `design` to allocate the samples on each batch and then the `spatial` function to randomly distribute the samples homogeneously by position within each batch.
-The samples are distributed on each position as with the `design` function, only returning the distribution with less differences between the statistics previously mentioned with the original distribution to the user of all the iterations tried.
-See below for an example:
+Samples are distributed on each position as with the `design` function, only returning to the user the distribution with less differences between the statistics previously mentioned with the original distribution of all the iterations tried.
+See below an example:
 

### 2021-08-25 paper.Rmd
Following some Joe comments on the text

@@
     df_print: paged
-date: "26st July 2021"
+date: "25th August 2021"
 affiliations:
@@
 The design of an experiment is critical to its success.
-Nonetheless, even with a correct design, the process up to the moment of measurement is critical.
-At any one of the several steps required from collection to measurement various errors and problems could affect the experimental results.
+Nonetheless, even when correctly designed, the process leading up to the moment of measuring a given variable is critical.
+At any one of the several steps, from sample collection to measurement of a variable, various errors and problems can affect the experimental results.
 Failure to take such variability into account can render an experiment inconclusive.
-*experDesign* provides tools to minimize the risk of inconclusive results by assigning samples to batches to minimize potential batch effects.
+*experDesign* provides tools to minimize the risk of inconclusive results by assigning samples to batches to reduce potential batch effects.
 
@@
 Sometimes, however, this is not possible.
-In this event, techniques to control such variations need to be applied.
-There are three techniques used to decrease the uncertainty of the unwanted variation: blocking, randomization and replication [@klaus2015].
+In such cases, techniques to control variations must to be applied.
+There are three methods used to decrease the uncertainty of the unwanted variation: blocking, randomization and replication [@klaus2015].
 
-**Blocking** is a method that groups samples that are equal according to one or more variables, allowing the estimation of the differences between each batch by comparing measurements within blocks.
+**Blocking** is a method that groups samples that are equal according to one or more variables, allowing the estimation of the differences between each batch by comparing measurements within the blocks.
 **Randomization** minimizes the variation in the measurements by randomly mixing the potential confounding variables.
 **Replication** increases the number of samples used in an experiment to better estimate the variation of the experiment.
-In some settings these techniques can be applied together to ensure the robustness of the study.
+In some settings these techniques can be applied together to enhance the robustness of the study.
 
-Between the designing of an experiment and the measurement of the samples, some samples might be lost, contaminated, or degradated below quality threshold.
+Between the designing of an experiment and the measurement of the samples, some samples might be lost, contaminated, or degraded below the quality threshold.
 In addition, experiments will occasionally need to be carried out in batches.
-Batches might be necessary for technical reasons, for example, the device cannot measure more than a number of samples at the same time.
-They can be also due to practical reasons, for instance, it may not be possible to obtain additional measurements in the field during the allotted time.
+The later might be needed for technical reasons; for example, the device cannot measure more than a given number of samples at the same time.
+Practical reasons can also be a factor; for instance, it may not be possible to obtain additional measurements in the field during the allotted time.
 
-This divergence from the original design might cause batch effects perturbing the analysis.
+This divergence from the original design might cause batch effects, thereby perturbing the analysis.
 There are several techniques to identify and assess batch effects when analyzing an already measured experiment [@leek2010].
 It would be better to avoid such batch effects before executing an experiment.
-By taking into account the de differences between the original design and the state before measurement is done, confounding effects could be minimized.
+By taking into account the differences between the original design and the state before the measurement is conducted, confounding effects can be minimized.
 
-To prevent the batch effect fom confounding the analysis after the initial design of the experiment, there are two options: randomization and replication.
-Randomization, consists on shuffling the samples to mix samples with different attributes, can help reduce variations across groups, while replication helps estimate the variation of the measurements or samples, thus increasing the precision of the estimates of the true value on the analysis.
+To prevent the batch effect from confounding the analysis after the initial design of the experiment, there are two options: randomization and replication.
+Randomization, consists of shuffling the samples in order to mix different attributes, which can help reduce variations across groups.
+In contrast, replication helps estimate the variation of the measurements or samples, thus increasing the precision of the estimates of the true value obtained by the analysis.
 Replications consist of increasing the number of measurements with similar attributes.
@@
 
-Randomization and replicates can be used to prevent batch effects to confound the analysis.
+Randomization and replication can be used to prevent batch effects that might confound the analysis.
 By examining how the variables are distributed across each batch, proper randomization can be ensured, thus minimizing batch effects.
@@
 
-There are some tools to minimize batch effects on the R language in multiple fields, and particularly for biological research [@rcoreteam2014].
+There are certain tools that can minimize batch effects on the R language in multiple fields, particularly for biological research [@rcoreteam2014].
 Here we briefly describe the currently available packages:
 
--   *OSAT*, at [Bioconductor](https://bioconductor.org/packages/OSAT/), first allocates the samples from each batch according to a variable; it then shuffles the samples from each batch in order to randomize the other variables [@yan2012].
-    This algorithm relies on categorical variables and cannot use numerical variables (e.g., age- or time-related) unless they are treated as categorical.
+-   [*OSAT*](https://bioconductor.org/packages/OSAT/), at Bioconductor, first allocates the samples from each batch according to a variable; it then shuffles the samples from each batch in order to randomize the other variables [@yan2012].
+    This algorithm relies on categorical variables and cannot use numerical variables (e.g., those that are age- or time-related) unless they are treated as categorical variables.
 
--   *anticlust*, at [CRAN](https://cran.r-project.org/package=anticlust), divides the samples into similar groups, ensuring similarity by enforcing heterogeneity within groups [@papenberg2020].
-    Conceptually it is similar to the clustering methods k-means.
+-   [*anticlust*](https://cran.r-project.org/package=anticlust), at CRAN, divides the samples into similar groups, ensuring similarity by enforcing heterogeneity within groups [@papenberg2020].
+    Conceptually it is similar to the clustering method k-means.
 
--   Recently, *Omixer*, a new package, has been made available at [Bioconductor](https://bioconductor.org/packages/Omixer/) [@sinke2021].
+-   Recently, [*Omixer*](https://bioconductor.org/packages/Omixer/), a new package, has been made available at Bioconductor [@sinke2021].
     It tests whether the random assignments made by it are homogeneous by transforming all variables to numeric values and using the Kendall's correlation when there are more than 5 samples; otherwise, it utilizes the Pearson's chi-squared test.
@@
 
-Current solutions to stratify samples to reduce and control batch effect do not work for all cases.
-They are either specialized to particular type of data, they omit some conditions that are often met or they only work on a specific subset of conditions.
-The new package ***experDesign*** works on all kind of data types and does not require a spatial distribution making it suitable for all kind of experiments.
-This package is intended for people looking for a quick and easy solution providing reasonable suggestions on how to distribute the samples for their analysis.
+Current solutions for stratifying samples to reduce and control batch effect do not work for all cases.
+They are either specialized to a particular type of data, they omit some conditions that are usually met, or they only work under a specific subset of conditions.
+The new package ***experDesign*** works with all data types and does not require a spatial distribution making it suitable for all kind of experiments.
+This package is intended for people needing a quick and easy solution that will provide reasonable suggestions on how to best distribute the samples for analysis.
 
-For completeness a description and comparison of the usage of the different software currently available on CRAN and Bioconductor is presented below.
-First, we start with some real data from a survey:
+For completeness a description and comparison of the usage of the different software packages currently available on CRAN and Bioconductor is presented below.
+First, we start with some real data obtained from a survey:
 
@@
 
-OSAT provides some template for plates that hold 2, 4, 8 Illumina BeadChip chips and have 24, 48, 96 wells and works for numeric and categorical variables:
+OSAT provides templates for plates that hold 2, 4, 8 Illumina BeadChip chips, having 24, 48 or 96 wells.
+Moreover, it works for both numeric and categorical variables:
 
-```{r osat, fig.cap="Samples distributions with OSAT."}
+```{r osat, fig.cap="Sample distributions with OSAT."}
 library("ggplot2")
@@
 
-OSAT returns one row less that the input provided.
-This row is deleted because it has an NA value on the Sex variable.
+OSAT returned one row less than the input provided.
+This row was deleted because it had an NA value for the Sex variable.
 
@@
 
-```{r anticlust, error=TRUE, fig.cap="Samples distributions with anticlust."}
+```{r anticlust, error=TRUE, fig.cap="Sample distributions with anticlust."}
 library("anticlust")
@@
 
-There is a bug on Omixer that prevents it to work unless specific conditions are meet.
-That prevents it comparing with the other tools using the same settings:
+There is a bug in the Omixer that prevents it from working unless specific conditions are meet.
+This precluded any comparisons of Omixer with other tools using the same settings:
 
@@
 
-The package ***experDesign*** provides the function `design` to arrange the samples into multiple batches such that the variable's distribution is homogeneous within each batch.
-Each batch is set to have some centrality and dispersion statistics to match as closely as possible the original input design data.
+The package ***experDesign*** provides the function `design` to arrange the samples into multiple batches such that a variable's distribution remains homogeneous within each batch.
+Each batch is set to have some centrality and dispersion statistics to match as closely as possible with the original input design data.
 The statistics used are the mean, the standard deviation, the median absolute deviation, variables with no value number, the entropy and the independence of the categorical variables.
-On each iteration if the random distribution of samples' statitics for each batch has fewer differences with the original distribution than the last stored sample distribution then it replaces it as the best sample distribution.
-At the end of the iterations the best sample distribution is returned to the user.
+With each iteration if the random distribution of the sample statistics for each batch has fewer differences vis-à-vis the original distribution than the last stored sample distribution then it replaces it as the best sample distribution.
+Upon completion of the iterations the best sample distribution is returned to the user.
 
-Users can follow the following flowchart to decide what function(s) they need to use:
+Users can examine the following flowchart to decide what function(s) they need to use:
 
@@
 
-If users want a design without replicates but batches have some spatial distribution, we need to use `design` to allocate the samples on each batch and then the `spatial` function to randomly distribute the samples homogeneously by position within each batch.
-Samples are distributed on each position as with the `design` function, only returning to the user the distribution with less differences between the statistics previously mentioned with the original distribution of all the iterations tried.
-See below an example:
+If users want a design without replicates but the batches have some spatial distribution, we must use `design` to allocate the samples on each batch, followed by the `spatial` function to randomly distribute the samples homogeneously by position within each batch.
+See the example below:
 
-```{r experDesign, fig.cap="Samples distributions with experDesign."}
+```{r experDesign, fig.cap="Sample distributions with experDesign."}
 library("experDesign")
@@
 In order to choose them from the available samples, the function `extreme_cases` is provided.
-For easier usage, the `replicates` function designs an experiment with the number of replicates per batch desired.
+For easier usage, the `replicates` function designs an experiment with the desired number of replicates per batch.
 The rest of the workflow is similar to the above example:
@@
 ***experDesign*** also provides several small utilities to make it easier to design the experiment in batches.
-For instance, a function called `sizes_batches` helps calculate the number of samples in order to distribute them across the batches required.
-Furthermore, `optimum_batches` calculates the minimal amount of batches required:
+For instance, a function called `sizes_batches` helps calculate the number of samples in order to distribute them across the required batches.
+Furthermore, `optimum_batches` calculates the minimal number of batches required:
 
@@
 
-In conclusion the ***experDesign*** can be a first and fast step to prepare for a batched experiment.
-It can use as many numeric and categorical variables as needed to stratify the experiment design on batches including spatial distributions.
+In conclusion ***experDesign*** offers a fast method for preparing a batched experiment.
+It can use as many numeric and categorical variables as needed to stratify the experimental design based on batches including spatial distributions.
 

### 2021-09-23 R/QC.R
Closes #34

@@ -89,6 +89,9 @@ qcSubset <- function(index, size, each = FALSE) {
 #' Check index distribution on batches
 #'
 #' Report the statistics for each subset and variable compared to the original.
+#'
+#' The closer the values are to 0, the less difference is with the original
+#' distribution, so it is a better randomization.
 #' @inheritParams design
 #' @inheritParams qcSubset
 #' @return A matrix with the differences with the original data.

### 2021-09-27 README.Rmd

@@
 [![R build status](https://github.com/llrs/experDesign/workflows/R-CMD-check/badge.svg)](https://github.com/llrs/experDesign/actions?workflow=R-CMD-check)
-[![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/llrs/experDesign?branch=master&svg=true)](https://ci.appveyor.com/project/llrs/experDesign)
-[![Travis build status](https://travis-ci.org/llrs/experDesign.svg?branch=master)](https://travis-ci.org/llrs/experDesign)
 [![Coverage status](https://codecov.io/gh/llrs/experDesign/branch/master/graph/badge.svg)](https://codecov.io/github/llrs/experDesign?branch=master)

### 2021-11-08 paper.Rmd
Remove long lines moves tatement of need below code examples
Update paper

@@
 output:
+  html_document:
+    df_print: paged
   word_document:
     keep_md: yes
+  pdf_document: default
   md_document:
     preserve_yaml: yes
-  pdf_document: default
-  html_document:
-    df_print: paged
 date: "25th August 2021"
@@
 
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(collapse = TRUE, warning = TRUE, error = TRUE, fig.width = 9)
+```{r setup}
+knitr::opts_chunk$set(collapse = TRUE, warning = TRUE, 
+                      error = 0,
+                      fig.width = 9)
 ```
@@
 
-# Statement of need
+For completeness a description and comparison of the usage of the different software packages currently available on CRAN and Bioconductor is presented below.
+First, we start with some real data obtained from a survey.
+This data set has three variables of interest; Sex, Smoke and Age are a mix of categorical and numeric variables.
 
-Current solutions for stratifying samples to reduce and control batch effect do not work for all cases.
-They are either specialized to a particular type of data, they omit some conditions that are usually met, or they only work under a specific subset of conditions.
-The new package ***experDesign*** works with all data types and does not require a spatial distribution making it suitable for all kind of experiments.
-This package is intended for people needing a quick and easy solution that will provide reasonable suggestions on how to best distribute the samples for analysis.
-
-For completeness a description and comparison of the usage of the different software packages currently available on CRAN and Bioconductor is presented below.
-First, we start with some real data obtained from a survey:
-
-```{r example-data}
+```{r example-data, include = FALSE}
 data(survey, package = "MASS")
@@
 
-This data set has three variables of interest; Sex, Smoke and Age are a mix of categorical and numeric variables.
-
 ## OSAT
@@
 
-```{r osat, fig.cap="Sample distributions with OSAT."}
+```{r osat, fig.cap="Sample, distributions with OSAT.", include = FALSE}
 library("ggplot2")
@@
 gs <- setup.sample(survey, optimal = VoI)
-gc <- setup.container(plate = IlluminaBeadChip96Plate, n = n_batch, batch = 'plates')
-gSetup <- create.optimized.setup(sample = gs, container = gc, nSim = iterations)
+gc <- setup.container(plate = IlluminaBeadChip96Plate, 
+                      n = n_batch, batch = 'plates')
+gSetup <- create.optimized.setup(sample = gs, container = gc, 
+                                 nSim = iterations)
 osat_report <- get.experiment.setup(gSetup)
@@
   labs(fill = "Batch")
-(osat_smoke/(osat_age + osat_sex + plot_layout(guides = 'collect'))) +
-  plot_annotation(title = "OSAT")
 ```
 
-OSAT returned one row less than the input provided.
-This row was deleted because it had an NA value for the Sex variable.
+OSAT might return less rows than the input provided because they might have NA value.
 
@@
 
-anticlust does not handle all the variables, it only accepts numeric variables:
+anticlust does not handle all the variables, it only accepts numeric variables.
 
-```{r anticlust, error=TRUE, fig.cap="Sample distributions with anticlust."}
+```{r anticlust, fig.cap="Sample, distributions with anticlust.", include = FALSE}
 library("anticlust")
@@
 )
-report_anticlust <- data.frame(Age = survey[, "Age"], batch = as.character(anticlust_index))
+report_anticlust <- data.frame(Age = survey[, "Age"], 
+                               batch = as.character(anticlust_index))
 library("ggplot2")
-ggplot(report_anticlust) + 
+anticlust <- ggplot(report_anticlust) + 
   geom_histogram(aes(Age, fill = batch)) +
   facet_wrap(~batch) +
-  labs(fill = "Batch", title = "anticlust")
+  labs(fill = "Batch")
 ```
@@
 There is a bug in the Omixer that prevents it from working unless specific conditions are meet.
-This precluded any comparisons of Omixer with other tools using the same settings:
+This precluded any comparisons of Omixer with other tools using the same settings.
 
-```{r Omixer, error=TRUE}
+```{r Omixer, include = FALSE}
 library("Omixer")
@@
 Omixer_index <- omixerRand(survey, sampleId = "ID", 
-                           block = "block", iterNum = 10, wells = size_subset, mask = rep(1, nrow(survey)),
+                           block = "block", iterNum = 10, wells = size_subset, 
+                           mask = rep(1, nrow(survey)),
                            plateNum = n_batch, randVars = VoI, techVars = NULL)
 Omixer_index <- omixerRand(survey, sampleId = "ID", 
-                           block = "block", iterNum = 10, wells = size_subset, mask = rep(0, nrow(survey)),
+                           block = "block", iterNum = 10, wells = size_subset, 
+                           mask = rep(0, nrow(survey)),
                            plateNum = n_batch, randVars = VoI, techVars = NULL)
 Omixer_index <- omixerRand(survey, sampleId = "ID", 
-                           block = "block", iterNum = 10, wells = size_subset, mask = rep(1, 288),
+                           block = "block", iterNum = 10, wells = size_subset, 
+                           mask = rep(1, 288),
                            plateNum = n_batch, randVars = VoI, techVars = NULL)
 Omixer_index <- omixerRand(survey, sampleId = "ID", 
-                           block = "block", iterNum = 10, wells = size_subset, mask = rep(0, 288),
+                           block = "block", iterNum = 10, wells = size_subset, 
+                           mask = rep(0, 288),
                            plateNum = n_batch, randVars = VoI, techVars = NULL)
@@
 ```
+
+```{r panel1, echo=FALSE, message=FALSE, warning=FALSE}
+(osat_smoke/(osat_age + osat_sex) / anticlust & plot_layout(guides = 'collect')
+```
+
+# Statement of need
+
+Current solutions for stratifying samples to reduce and control batch effect do not work for all cases.
+They are either specialized to a particular type of data, they omit some conditions that are usually met, or they only work under a specific subset of conditions.
+The new package ***experDesign*** works with all data types and does not require a spatial distribution making it suitable for all kind of experiments.
+This package is intended for people needing a quick and easy solution that will provide reasonable suggestions on how to best distribute the samples for analysis.
 
@@
 If users want a design without replicates but the batches have some spatial distribution, we must use `design` to allocate the samples on each batch, followed by the `spatial` function to randomly distribute the samples homogeneously by position within each batch.
-See the example below:
+See the example in `inspect` and the vignette:
 
-```{r experDesign, fig.cap="Sample distributions with experDesign."}
+```{r experDesign, echo=FALSE, fig.cap="Sample, distributions with anticlust.", warning=FALSE, message=FALSE}
 library("experDesign")
@@
                   iterations = iterations, name = "")
-experDesign_index_spatial <- spatial(index = experDesign_index, pheno = survey[, VoI], 
-                                     iterations = iterations, rows = LETTERS[1:8], columns = 1:12)
+experDesign_index_spatial <- spatial(index = experDesign_index, 
+                                     pheno = survey[, VoI], 
+                                     iterations = iterations, 
+                                     rows = LETTERS[1:8], 
+                                     columns = 1:12)
 experDesign_report <- inspect(experDesign_index, survey[, VoI])
@@
   labs(fill = "Batch")
-(experDesign_smoke/(experDesign_age + experDesign_sex + plot_layout(guides = 'collect'))) + plot_annotation(title = "experDesign")
+(experDesign_smoke/(experDesign_age + 
+                      experDesign_sex )) +
+  plot_annotation(title = "experDesign") +
+  plot_layout(guides = 'collect')
 ```
 
-The statistics of the index can be checked for multiple statistics, as shown in the following code:
+The statistics of the index can be checked for multiple statistics, as shown on the help pages of `evaluate_na`, `evaluate_entropy`, `evaluate_mad`, `evaluate_sd` and `evaluate_mean`.
 
-```{r evaluate}
+```{r evaluate, include = FALSE}
 evaluate_na(experDesign_index, survey[, VoI])
@@
 
-We can also compare our results with the original distribution:
+We can also compare our results with the original distribution via `evaluate_orig`.
 
-```{r original}
+```{r original, include = FALSE}
 evaluate_orig(survey[, VoI])
@@
 For easier usage, the `replicates` function designs an experiment with the desired number of replicates per batch.
-The rest of the workflow is similar to the above example:
 
-```{r controls}
+```{r controls, include = FALSE}
 extreme_cases(survey[, VoI], size = 5)
@@
                                iterations = iterations)
-spatial_replicate_index <- spatial(index = replicates_index, pheno = survey[, VoI], 
+spatial_replicate_index <- spatial(index = replicates_index, 
+                                   pheno = survey[, VoI], 
                                    rows = LETTERS[1:8], columns = 1:12)
@@
 
-```{r show-batches}
+```{r show-batches, include = FALSE}
 optimum_batches(size_data = 250, size_subset = 96)

### 2021-11-10 paper.Rmd
Update plots and text accordingly

@@
 output:
-  html_document:
-    df_print: paged
+  md_document:
+    preserve_yaml: yes
   word_document:
@@
   pdf_document: default
-  md_document:
-    preserve_yaml: yes
-date: "25th August 2021"
+  html_document:
+    df_print: paged
+date: "11th November 2021"
 affiliations:
@@
 
-```{r setup}
-knitr::opts_chunk$set(collapse = TRUE, warning = TRUE, 
+```{r setup, include = FALSE}
+knitr::opts_chunk$set(
 	error = 0,
-                      fig.width = 9)
+	fig.width = 9,
+	warning = TRUE,
+	collapse = TRUE
+)
 ```
@@
 OSAT provides templates for plates that hold 2, 4, 8 Illumina BeadChip chips, having 24, 48 or 96 wells.
-Moreover, it works for both numeric and categorical variables:
+Moreover, it works for both numeric and categorical variables but OSAT might return less rows than the input provided because they might have NA value.
 
@@
 
-OSAT might return less rows than the input provided because they might have NA value.
-
 ## anticlust
 
-anticlust does not handle all the variables, it only accepts numeric variables.
+anticlust does not handle all types of variables, it only accepts numeric variables.
 
@@
 ```
+These methods results on this distributions:
 
-```{r panel1, echo=FALSE, message=FALSE, warning=FALSE}
-(osat_smoke/(osat_age + osat_sex) / anticlust & plot_layout(guides = 'collect')
+```{r panel1, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Figure A: distribution of  samples according to *OSAT*, Figure B according to *anticlust* on the variables used (1: Smoke, 2: Age, 3: Sex)."}
+osat_plot <- osat_smoke / (osat_age + osat_sex)
+osat_plot /anticlust + plot_layout(guides = 'collect') +
+  plot_annotation(tag_levels = list(c("A.1", "A.2", "A.3", "B.2")),
+                  title = "Existing solutions",
+                  subtitle = "OSAT and anticlust solutions") &
+  theme(plot.tag.position = c(0, 1),
+        plot.tag = element_text(size = 8, hjust = 0, vjust = 0))
 ```
+
+On Figure 1, we can see that the distributions is fairly similar between these two packages.
 
@@
 experDesign_report <- inspect(experDesign_index, survey[, VoI])
-nrow(experDesign_report)
+# nrow(experDesign_report)
 experDesign_age <- ggplot(experDesign_report) + 
@@
 The statistics of the index can be checked for multiple statistics, as shown on the help pages of `evaluate_na`, `evaluate_entropy`, `evaluate_mad`, `evaluate_sd` and `evaluate_mean`.
+We can also compare our results with the original distribution via `evaluate_orig`.
 
@@
 ```
-
-We can also compare our results with the original distribution via `evaluate_orig`.
 

### 2021-11-10 paper.Rmd
Fix legend and add a sentence about examples

@@
 
-```{r experDesign, echo=FALSE, fig.cap="Sample, distributions with anticlust.", warning=FALSE, message=FALSE}
+```{r experDesign, echo=FALSE, fig.cap="Sample, distributions with ***experDesign***.", warning=FALSE, message=FALSE}
 library("experDesign")
@@
 For instance, a function called `sizes_batches` helps calculate the number of samples in order to distribute them across the required batches.
-Furthermore, `optimum_batches` calculates the minimal number of batches required:
+Furthermore, `optimum_batches` calculates the minimal number of batches required.
+Examples of all this methods can be found on the manual page of each function and on the vignette.
 

### 2021-11-18 paper.Rmd
Remove figure 1, move text and thank editor and reviewers.

@@
 -   [*OSAT*](https://bioconductor.org/packages/OSAT/), at Bioconductor, first allocates the samples from each batch according to a variable; it then shuffles the samples from each batch in order to randomize the other variables [@yan2012].
-    This algorithm relies on categorical variables and cannot use numerical variables (e.g., those that are age- or time-related) unless they are treated as categorical variables.
+    This algorithm relies on categorical variables and cannot use numerical variables (e.g., those that are age- or time-related) unless they are treated as categorical variables.\
+    OSAT provides templates for plates that hold 2, 4, 8 Illumina BeadChip chips, having 24, 48 or 96 wells.
+    Moreover, it works for both numeric and categorical variables but OSAT might return less rows than the input provided because they might have NA value.
 
 -   [*anticlust*](https://cran.r-project.org/package=anticlust), at CRAN, divides the samples into similar groups, ensuring similarity by enforcing heterogeneity within groups [@papenberg2020].
-    Conceptually it is similar to the clustering method k-means.
+    Conceptually it is similar to the clustering method k-means.\
+    anticlust does not handle all types of variables, it only accepts numeric variables.
 
 -   Recently, [*Omixer*](https://bioconductor.org/packages/Omixer/), a new package, has been made available at Bioconductor [@sinke2021].
-    It tests whether the random assignments made by it are homogeneous by transforming all variables to numeric values and using the Kendall's correlation when there are more than 5 samples; otherwise, it utilizes the Pearson's chi-squared test.
+    It tests whether the random assignments made by it are homogeneous by transforming all variables to numeric values and using the Kendall's correlation when there are more than 5 samples; otherwise, it utilizes the Pearson's chi-squared test.\
+    There is a bug in the Omixer that prevents it from working unless specific conditions are meet.
+    This precluded any comparisons of Omixer with other tools using the same settings.
 
@@
 ```
-
-## OSAT
-
-OSAT provides templates for plates that hold 2, 4, 8 Illumina BeadChip chips, having 24, 48 or 96 wells.
-Moreover, it works for both numeric and categorical variables but OSAT might return less rows than the input provided because they might have NA value.
 
@@
 
-## anticlust
-
-anticlust does not handle all types of variables, it only accepts numeric variables.
-
 ```{r anticlust, fig.cap="Sample, distributions with anticlust.", include = FALSE}
@@
 
-## Omixer
-
-There is a bug in the Omixer that prevents it from working unless specific conditions are meet.
-This precluded any comparisons of Omixer with other tools using the same settings.
-
 ```{r Omixer, include = FALSE}
@@
 ```
-These methods results on this distributions:
-
-```{r panel1, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Figure A: distribution of  samples according to *OSAT*, Figure B according to *anticlust* on the variables used (1: Smoke, 2: Age, 3: Sex)."}
-osat_plot <- osat_smoke / (osat_age + osat_sex)
-osat_plot /anticlust + plot_layout(guides = 'collect') +
-  plot_annotation(tag_levels = list(c("A.1", "A.2", "A.3", "B.2")),
-                  title = "Existing solutions",
-                  subtitle = "OSAT and anticlust solutions") &
-  theme(plot.tag.position = c(0, 1),
-        plot.tag = element_text(size = 8, hjust = 0, vjust = 0))
-```
-
-On Figure 1, we can see that the distributions is fairly similar between these two packages.
 
@@
 
-```{r experDesign, echo=FALSE, fig.cap="Sample, distributions with ***experDesign***.", warning=FALSE, message=FALSE}
+```{r experDesign, echo=FALSE, fig.cap="Example of distributions generated with ***experDesign***.", warning=FALSE, message=FALSE}
 library("experDesign")
@@
 ```
+
+On Figure 1 one can see that the distribution index generated by experDesign uses both numeric and categorical variables keeping also the samples with missing (NA) information.
 
@@
 
-We are grateful to Joe Moore for English-language assistance.
+We are grateful to Joe Moore for English-language assistance and we would like to thank reviewers and the editor for taking the time and effort necessary to review the manuscript.
 

### 2021-11-22 paper.Rmd

@@
 They are either specialized to a particular type of data, they omit some conditions that are usually met, or they only work under a specific subset of conditions.
-The new package ***experDesign*** works with all data types and does not require a spatial distribution making it suitable for all kind of experiments.
+The new package [***experDesign***](https://cran.r-project.org/package=experDesign "experDesign on CRAN") works with all data types and does not require a spatial distribution making it suitable for all kind of experiments.
 This package is intended for people needing a quick and easy solution that will provide reasonable suggestions on how to best distribute the samples for analysis.
@@
 
-In conclusion ***experDesign*** offers a fast method for preparing a batched experiment.
+In conclusion [***experDesign***](https://cran.r-project.org/package=experDesign "experDesign pacakge") offers a fast method for preparing a batched experiment.
 It can use as many numeric and categorical variables as needed to stratify the experimental design based on batches including spatial distributions.

### 2021-11-22 README.Rmd
Increment version number to 0.1.1

@@
 [![Project Status: Active - The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
+[![DOI](https://zenodo.org/badge/142569201.svg)](https://zenodo.org/badge/latestdoi/142569201)
+
 <!-- badges: end -->

### 2022-12-23 R/designer.R
Fix examples to avoid warnings

@@ -24,6 +24,12 @@ design <- function(pheno, size_subset, omit = NULL, iterations = 500,
   stopifnot(length(dim(pheno)) == 2)
   stopifnot(is.numeric(iterations) && is.finite(iterations))
   stopifnot(is.character(name))
+  .design(pheno, size_subset, omit, iterations, name, check = TRUE)
+}
+
+
+.design <- function(pheno, size_subset, omit = NULL, iterations = 500,
+                    name = "SubSet", check = FALSE) {
   opt <- Inf
 
   # Calculate batches
@@ -39,8 +45,8 @@ design <- function(pheno, size_subset, omit = NULL, iterations = 500,
   }
 
   pheno_o <- omit(pheno, omit)
-  if (!.check_data(pheno_o)) {
-    warning("There might be some problems with the data use check_data().")
+  if (check && !.check_data(pheno_o)) {
+    warning("There might be some problems with the data use check_data().", call. = FALSE)
   }
 
   original_pheno <- .evaluate_orig(pheno_o)
@@ -49,7 +55,7 @@ design <- function(pheno, size_subset, omit = NULL, iterations = 500,
   # Find the numeric values
   dates <- vapply(pheno_o, function(x){methods::is(x, "Date")}, logical(1L))
   if (any(dates)) {
-    warning("The dates will be treat as categories")
+    warning("The dates will be treat as categories", call. = FALSE)
   }
 
   num <- is_num(pheno_o)
@@ -78,6 +84,7 @@ design <- function(pheno, size_subset, omit = NULL, iterations = 500,
   }
   val
 }
+
 
 #' Design a batch experiment with experimental controls
 #'
@@ -111,7 +118,7 @@ replicates <- function(pheno, size_subset, controls, omit = NULL,
   }
   size_subset <- size_subset - controls
   values <- extreme_cases(pheno = pheno, size = controls, omit = omit)
-  batches <- design(pheno, size_subset, omit = omit, iterations = iterations)
+  batches <- .design(pheno, size_subset, omit = omit, iterations = iterations)
 
   for (b in seq_along(batches)) {
     val <- c(batches[[b]], values[!values %in% batches[[b]]])

### 2022-12-23 R/evaluate_category.R

@@ -96,7 +96,7 @@ check_data <- function(pheno, na.omit = FALSE) {
   .check_data(pheno = pheno, na.omit = na.omit, verbose = TRUE)
 }
 
-.check_data <- function(pheno, na.omit, verbose = FALSE) {
+.check_data <- function(pheno, na.omit = FALSE, verbose = FALSE) {
 
   data_status <- TRUE
   num <- is_num(pheno)
@@ -123,7 +123,7 @@ check_data <- function(pheno, na.omit = FALSE) {
   nas <- lapply(pheno_o, function(x){which(is.na(x))})
   if (any(lengths(nas) >= 1)) {
     if (verbose) {
-      warning("Some values are missing", call. = FALSE)
+      warning("Some values are missing.", call. = FALSE)
     }
     data_status <- if (!na.omit) FALSE
   }

### 2022-12-23 R/reporting.R

@@ -45,7 +45,8 @@ inspect <- function(i, pheno, omit = NULL, index_name = "batch") {
 #' @examples
 #' data(survey, package = "MASS")
 #' columns <- c("Sex", "Age", "Smoke")
-#' index <- design(pheno = survey[, columns], size_subset = 70,
+#' nas <- c(137, 70) # Omit rows with NA to avoid warnings in design
+#' index <- design(pheno = survey[-nas, columns], size_subset = 70,
 #'                 iterations = 10)
 #' batches <- inspect(index, survey[, columns])
 #' distribution(batches, "Sex")

### 2022-12-23 README.Rmd
Update documentation

--- 2021-09-27 README.Rmd
@@
 [![CRAN status](https://www.r-pkg.org/badges/version/experDesign)](https://CRAN.R-project.org/package=experDesign)
-[![R build status](https://github.com/llrs/experDesign/workflows/R-CMD-check/badge.svg)](https://github.com/llrs/experDesign/actions?workflow=R-CMD-check)
+[![R-CMD-check](https://github.com/llrs/experDesign/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/llrs/experDesign/actions/workflows/R-CMD-check.yaml)
 [![Coverage status](https://codecov.io/gh/llrs/experDesign/branch/master/graph/badge.svg)](https://codecov.io/github/llrs/experDesign?branch=master)
@@
 [![Project Status: Active - The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
+[![JOSS](https://joss.theoj.org/papers/10.21105/joss.03358/status.svg)](https://doi.org/10.21105/joss.03358) 
+[![DOI](https://zenodo.org/badge/142569201.svg)](https://zenodo.org/badge/latestdoi/142569201)
+
 <!-- badges: end -->
@@
 
+* `check_data()`: Check if there are any problems with the data.
 * `design()`: Randomize the samples according to their variables.
@@
 The dataset has numeric, categorical values and some `NA`'s value.
+
+We can check some issues from an experimental point of view via `check_data()`:
+
+```{r check, warning=TRUE}
+check_data(survey)
+```
+
+As you can see with the warnings we get a collections of problems.
+In general, try to have at least 3 replicates for each condition and try to have all the data of each variable.
 

### 2022-12-23 R/QC.R
TODO: Fix test empty_res and remove all the insert functions

@@ -18,8 +18,8 @@ extreme_cases <- function(pheno, size, omit = NULL, iterations = 500) {
 
   # Calculate batches
   pheno_o <- omit(pheno, omit)
-
-  original_pheno <- .evaluate_orig(pheno_o)
+  num <- is_num(pheno_o)
+  original_pheno <- .evaluate_orig(pheno_o, num = num)
 
   # Find the numeric values
   dates <- vapply(pheno_o, function(x){methods::is(x, "Date")}, logical(1L))
@@ -33,7 +33,7 @@ extreme_cases <- function(pheno, size, omit = NULL, iterations = 500) {
   for (x in seq_len(iterations)) {
     i <- create_index(nSamples, size_batches, 1)
 
-    subsets <- evaluate_index(i, pheno_o)
+    subsets <- .evaluate_index(i, pheno_o, num)
     # Evaluate the differences between the subsets and the originals
     differences <- drop(abs(sweep(subsets, c(1, 2), original_pheno)))
     differences <- differences[-c(1, 4), ]
@@ -115,10 +115,10 @@ check_index <- function(pheno, index, omit = NULL) {
   # check this on evaluate_index
   eval_n <- ifelse(num, 4, 3)
 
-  original_pheno <- .evaluate_orig(pheno_o)
+  original_pheno <- .evaluate_orig(pheno_o, num)
   original_pheno["na", ] <- original_pheno["na", ]/batches
 
-  subsets <- evaluate_index(index, pheno_o)
+  subsets <- .evaluate_index(index, pheno_o, num)
   # Evaluate the differences between the subsets and the originals
   differences <- abs(sweep(subsets, c(1, 2), original_pheno))
   # Add the independence of the categorical values

### 2022-12-23 R/designer.R

@@ -49,7 +49,8 @@ design <- function(pheno, size_subset, omit = NULL, iterations = 500,
     warning("There might be some problems with the data use check_data().", call. = FALSE)
   }
 
-  original_pheno <- .evaluate_orig(pheno_o)
+  num <- is_num(pheno_o)
+  original_pheno <- .evaluate_orig(pheno_o, num)
   original_pheno["na", ] <- original_pheno["na", ]/batches
 
   # Find the numeric values
@@ -58,7 +59,6 @@ design <- function(pheno, size_subset, omit = NULL, iterations = 500,
     warning("The dates will be treat as categories", call. = FALSE)
   }
 
-  num <- is_num(pheno_o)
   # Numbers are evaluated 4 times, and categories only 3
   # check this on evaluate_index
   eval_n <- ifelse(num, 4, 3)
@@ -66,7 +66,7 @@ design <- function(pheno, size_subset, omit = NULL, iterations = 500,
   for (x in seq_len(iterations)) {
     i <- create_index(size_data, size_batches, batches, name = name)
 
-    subsets <- evaluate_index(i, pheno_o)
+    subsets <- .evaluate_index(i, pheno_o, num)
     # Evaluate the differences between the subsets and the originals
     differences <- abs(sweep(subsets, c(1, 2), original_pheno))
     # Add the independence of the categorical values

### 2022-12-23 R/evaluate.R

@@ -33,37 +33,16 @@ evaluate_orig <- function(pheno) {
   if (!.check_data(pheno)) {
     warning("There might be some problems with the data use check_data().")
   }
-  .evaluate_orig(pheno)
+  num <- is_num(pheno)
+  .evaluate_orig(pheno, num)
 }
 
-.evaluate_orig <- function(pheno) {
+.evaluate_orig <- function(pheno, num) {
   stopifnot(!is.null(colnames(pheno)))
-  original <- summary_num(pheno)
+  original <- empty_res(pheno, num)
+  original <- insert(original, colSums(is.na(pheno)), "na")
 
-  na_orig <- colSums(is.na(pheno))
-  original <- insert(original, na_orig, "na")
-
-  num <- is_num(pheno)
-
-  # Numeric data
-  if (sum(num) >= 1) {
-    pheno_num <- pheno[, num, drop = FALSE]
-
-    subset_num <- apply(pheno_num, 2, function(y) {
-      c("sd" = sd(y, na.rm = TRUE),
-        "mean" = mean(y, na.rm = TRUE),
-        "mad" = mad(y, na.rm = TRUE))
-    })
-    original <- insert(original, subset_num, c("sd", "mean", "mad"))
-  }
-  # Categorical data
-  if (sum(!num) >= 1) {
-    pheno_cat <- pheno[, !num, drop = FALSE]
-    entropy_orig <- apply(pheno_cat, 2, entropy)
-    original <- insert(original, entropy_orig, "entropy")
-  }
-
-  original
+  ev_subset(x = seq_len(nrow(pheno)), pheno = pheno, numeric = num, diff = original)
 }
 
 #' Evaluates a data.frame
@@ -83,42 +62,55 @@ evaluate_orig <- function(pheno) {
 #' ev_index <- evaluate_index(index, survey[, c("Sex", "Smoke")])
 #' ev_index["entropy", , ]
 evaluate_index <- function(i, pheno) {
-  if (!(is.matrix(pheno) || is.data.frame(pheno))) {
-    stop("Please provide a matrix or a data.frame")
+  if (!.check_data(pheno)) {
+    warning("There might be some problems with the data use check_data().", call. = FALSE)
   }
-  num <- is_num(pheno)
+  .evaluate_index(i, pheno, is_num(pheno))
 
-  diff <- summary_num(pheno)
-  ev_subset <- function(x){
+}
 
[…]

### 2022-12-23 R/spatial.R

@@ -53,7 +53,8 @@ spatial <- function(index, pheno, omit = NULL, remove_positions = NULL, rows = L
   batches <- length(index)
   pheno_o <- omit(pheno, omit)
 
-  original_pheno <- .evaluate_orig(pheno_o)
+  num <- is_num(pheno_o)
+  original_pheno <- .evaluate_orig(pheno_o, num)
   original_pheno["na", ] <- original_pheno["na", ]/batches
 
   # Find the numeric values
@@ -62,7 +63,6 @@ spatial <- function(index, pheno, omit = NULL, remove_positions = NULL, rows = L
     warning("The dates will be treat as categories")
   }
 
-  num <- is_num(pheno_o)
   # Numbers are evaluated 4 times, and categories only 3
   # check this on evaluate_index
   eval_n <- ifelse(num, 4, 3)
@@ -79,7 +79,7 @@ spatial <- function(index, pheno, omit = NULL, remove_positions = NULL, rows = L
   for (x in seq_len(iterations)) {
     i <- create_index(size_data, size_batches, batches, name = position)
 
-    subsets <- evaluate_index(i, pheno_o)
+    subsets <- .evaluate_index(i, pheno_o, num)
     # Evaluate the differences between the subsets and the originals
     differences <- abs(sweep(subsets, c(1, 2), original_pheno))
     # Add the independence of the categorical values

### 2022-12-23 R/utils.R

@@ -45,7 +45,7 @@ position_name <- function(rows, columns) {
   positions
 }
 
-summary_num <- function(pheno) {
+empty_res <- function(pheno, num) {
   if (is.null(ncol(pheno))) {
     ncol <- 1
     column <- "variable"
@@ -53,8 +53,14 @@ summary_num <- function(pheno) {
     ncol <- ncol(pheno)
     column <- colnames(pheno)
   }
+  if (sum(!num) > 2) {
+    ncol <- ncol +1
+    column <- c(column, "mix_cat")
+  }
+
+
   diff <- matrix(0, ncol = ncol, nrow = 5)
-  rownames(diff) <- c("mean", "sd", "mad", "na", "entropy")
+  rownames(diff) <- c("mean", "sd", "mad", "na", "entropy", "independence")
   colnames(diff) <- column
   diff
 }

### 2022-12-23 tests/testthat/test-empty_res.R

@@ -1,0 +1,4 @@
+test_that("empty_res works", {
+  d <- data.frame(a = letters)
+  empty_res(d, is_num(d))
+})

### 2023-01-13 R/follow_up.R
Initial work in follow_up functions
[Skip CI]

@@ -1,0 +1,116 @@
+#' Follow up experiments
+#'
+#' If an experiment was carried out with some samples and you want to continue
+#' with some other samples later on.
+#' @param original A `data.frame` with the information of the samples used originally.
+#' @param follow_up A `data.frame` with the information of the new samples.
+#' @param old_new Name of the column where the batch status will be stored. If
+#' it matches the name of a column in original it will be used to find previous
+#' batches.
+#' @seealso follow_up2()
+#' @return
+#' @export
+#'
+#' @examples
+follow_up <- function(original, follow_up, old_new = "batch") {
+  stopifnot(is.character(old_new))
+  if (!.check_data(original)) {
+    warning("There might be some problems with the original data use check_data().")
+  }
+
+  stopifnot(!is.data.frame(follow_up))
+  match_columns <- intersect(colnames(original), colnames(follow_up))
+
+  if (length(match_columns) == 0) {
+    stop("No shared column between the two data.frames")
+  }
+  if (old_new %in% colnames(original)) {
+    warning("There is already a follow up study, use follow_up2.")
+  }
+
+  original[[old_new]] <- "old"
+  follow_up[[old_new]] <- "new"
+
+
+
+}
+
+#' Follow up experiments in batches
+#'
+#' Design experiment with all the data new and old together.
+#'
+#' If the `batch_column` is empty the samples are considered new.
+#' If the `size_subset` is missing, it will be estimated from the previous batch
+#' Similarly, iterations and name will be guessed or inferred from the samples.
+#' @param all_data A `data.frame` with all the data about the samples.
+#' Each row is a sample.
+#' @param batch_column The name of the column of `all_data` with the batches used.
+#' If NA it is interpreted as a new data, if not empty it is considered a batch.
+#' @inheritDotParams design -pheno
+#' @seealso `follow_up()`
+#' @return A `data.frame` with the `batch_column` filled with the new batches needed.
+#' @export
+#'
+#' @examples
+#' data(survey, package = "MASS")
+#' # Create the first batch
+#' first_batch_n <- 118
+#' variables <- c("Sex", "Smoke", "Age")
+#' survey1 <- survey[seq_len(first_batch_n), variables]
+#' index1 <- design(survey1, size_subset = 50, iterations = 50)
[…]

### 2023-01-13 R/indexing.R

@@ -34,19 +34,21 @@ create_subset <- function(size_data, size_subset = NULL, n = NULL, name = "SubSe
 
 # The workhorse function without any check
 create_index <- function(size_data, size_batches, n, name = "SubSet") {
+  # The size of each batch
+  i <- distribute_samples(size_data, size_batches)
+  names(i) <- id2batch_names(name, n)
+  i
+}
 
+id2batch_names <- function(name, n) {
   if (length(name) != 1 && length(name) != n) {
     stop("Provide a single character or a vector the same size of the batches.",
          call. = FALSE)
-  } else if (length(name) == 1) {
+  }
+  if (length(name) == 1) {
     name <- paste0(name, seq_len(n))
   }
-
-  # The size of each batch
-  i <- distribute_samples(size_data, size_batches)
-  names(i) <- name
-  i
-
+  name
 }
 
 distribute_samples <- function(size_data, size_subsets) {
@@ -71,11 +73,20 @@ distribute_samples <- function(size_data, size_subsets) {
 #' @examples
 #' plates <- c("P1", "P2", "P1", "P2", "P2", "P3", "P1", "P3", "P1", "P1")
 #' use_index(plates)
-use_index <- function(x){
-  split(seq_along(x), x)
+use_index <- function(x) {
+  stopifnot(is.character(x))
+  if (anyNA(x)) {
+    warning("NAs present in the index. Some samples weren't assigned to a batch?")
+  }
+  .use_index(x)
 }
 
-#
+.use_index <- function(x) {
+  factors <- x
+  factors[is.na(x)] <- "NA"
+  split(seq_along(x), factors)
+}
+
 #' Name the batch
 #'
 #' Given an index return the name of the batches the samples are in

### 2023-01-13 tests/testthat/test-follow_up.R

@@ -1,0 +1,16 @@
+test_that("follow_up works", {
+  data(survey, package = "MASS")
+  survey$batch[1:118] <- "old"
+
+})
+test_that("follow_up2 works", {
+  data(survey, package = "MASS")
+  old_n <- 118
+  bn <- batch_names(create_subset(old_n, 20))
+  survey$batch <- c(bn, rep(NA, nrow(survey) - old_n))
+  follow_up2(survey)
+  survey$batch <- c(rep("old", old_n), rep(NA, nrow(survey) - old_n))
+  follow_up2(survey)
+  survey$batch <- ifelse(survey$Clap == "Right", "old", "new")
+  expect_error(follow_up2(survey))
+})

### 2023-01-13 README.Rmd
Reduce execution of the vignette

@@
 [![CRAN status](https://www.r-pkg.org/badges/version/experDesign)](https://CRAN.R-project.org/package=experDesign)
+[![cran checks](https://badges.cranchecks.info/worst/experDesign.svg)](https://cran.r-project.org/web/checks/check_results_experDesign.html)
 [![R-CMD-check](https://github.com/llrs/experDesign/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/llrs/experDesign/actions/workflows/R-CMD-check.yaml)
@@
 
-The goal of experDesign is to help you decide which samples go in which batch, 
-reducing the potential batch bias before performing an experiment.
-It provides three main functions :
+The goal of experDesign is to help you manage your samples before an experiment but after they are collected.
+For example, checking for common problems in the data, and help reduce or even prevent batch bias before performing an experiment, or measure once it is already done
+It provides four main functions:
 
@@
 * `design()`: Randomize the samples according to their variables.
-* `replicates()`: Selects some samples for replicates and randomizes the samples.
+* `replicates()`: Selects some samples for replicates and randomizes the samples (highly recommended).
 * `spatial()`: Randomize the samples on a spatial grid.
@@
 
+### Checking initial data
+
 We can check some issues from an experimental point of view via `check_data()`:
@@
 
-# Picking samples for each batch
+### Picking samples for each batch
 
 Imagine that we can only work in groups of 70, and we want to randomize by Sex, 
-Smoke, Age, and by writing hand.  
-There are `r choose(237, 70)` combinations some of them would be have in a 
-single experiment all the right handed students. We could measure all these combinations
-but we can try to find an optimum value.
+Smoke, Age, and by hand.  
+There are `r choose(237, 70)` combinations of samples per batch in a 
+this experiment. 
+However, in some of those combinations all the right handed students are in the same batch making it impossible to compare the right handed students with the others and draw conclusions from it.
+
+We could check all the combinations to select those that allow us to do this comparison.
+But as this would be too long with `experDesign` we can try to find the combination with the best design by comparing each combination with the original according to multiple statistics.
 
@@
 
+# Set a seed for reproducibility
+set.seed(87732135)
 # Looking for groups at most of 70 samples.
-index <- design(pheno = survey, size_subset = 70, omit = omit)
+index <- design(pheno = survey, size_subset = 70, omit = omit, iterations = 100)
 index
@@
 
-We can transform then into a vector to append to the file or to pass to the lab mate with:
+We can transform then into a vector to append to the file or to pass to a colleague with:
 
@@
 head(batch_names(index))
+# Or via inspect() to keep it in a matrix format:
+head(inspect(index, survey[, keep]))
 ```

### 2023-01-27 R/evaluate_category.R
Check pairwise problems when there are more than one category

@@ -119,17 +119,19 @@ check_data <- function(pheno, na.omit = FALSE) {
     return(data_status)
   }
 
-  pairwise_colusion <- combn(names(pheno)[cat], 2, FUN = function(x) {
-    y <- table(pheno[, x], useNA = if (!na.omit) "ifany")
-    any(y == 0)
-  })
+  if (sum(cat) >= 2) {
+    pairwise_colusion <- combn(names(pheno)[cat], 2, FUN = function(x) {
+      y <- table(pheno[, x], useNA = if (!na.omit) "ifany")
+      any(y == 0)
+    })
 
-  if (any(pairwise_colusion)) {
-    if (isTRUE(verbose)) {
-      warning("Two categorical variables don't have all the combinations.",
-              call. = FALSE)
+    if (any(pairwise_colusion)) {
+      if (isTRUE(verbose)) {
+        warning("Two categorical variables don't have all combinations.",
+                call. = FALSE)
+      }
+      data_status <- FALSE
     }
-    data_status <- FALSE
   }
 
   pheno_o <- droplevels(pheno[, cat, drop = FALSE])

### 2023-01-27 R/follow_up.R
Improve follow_up

@@ -7,18 +7,23 @@
 #' @param old_new Name of the column where the batch status will be stored. If
 #' it matches the name of a column in original it will be used to find previous
 #' batches.
+#' @inheritParams design
 #' @seealso follow_up2()
 #' @return
 #' @export
-#'
 #' @examples
-follow_up <- function(original, follow_up, old_new = "batch") {
+#' data(survey, package = "MASS")
+#' survey1 <- survey[1:118, ]
+#' survey2 <- survey[119:nrow(survey), ]
+#' fu <- follow_up(survey1, survey2, iterations = 10)
+follow_up <- function(original, follow_up, old_new = "batch", iterations = 500) {
   stopifnot(is.character(old_new) & length(old_new) == 1)
+  stopifnot(is.data.frame(original))
   if (!.check_data(original)) {
     warning("There might be some problems with the original data use check_data().")
   }
 
-  stopifnot(!is.data.frame(follow_up))
+  stopifnot(is.data.frame(follow_up))
   match_columns <- intersect(colnames(original), colnames(follow_up))
 
   if (length(match_columns) == 0) {

### 2023-01-27 tests/testthat/test-follow_up.R

@@ -2,7 +2,7 @@ test_that("follow_up works", {
   data(survey, package = "MASS")
   survey1 <- survey[1:118, ]
   survey2 <- survey[119:nrow(survey), ]
-  fu <- follow_up(surve1, survey2)
+  expect_warning(fu <- follow_up(survey1, survey2, iterations = 10))
 })
 
 test_that("follow_up2 works", {
@@ -10,21 +10,21 @@ test_that("follow_up2 works", {
   old_n <- 118
   # old vs new
   survey$batch <- c(rep("old", old_n), rep(NA, nrow(survey) - old_n))
-  follow_up2(survey)
+  expect_snapshot(fu1 <- follow_up2(survey))
 
   # old vs new with counfounding effects
   survey$batch <- ifelse(survey$Clap %in% "Right", "old", NA)
-  expect_error(follow_up2(survey))
+  expect_error(expect_warning(expect_warning(expect_warning(follow_up2(survey, iterations = 10)))))
 
   # old with batches vs new
   bn <- batch_names(create_subset(old_n, 20))
   survey$batch <- c(bn, rep(NA, nrow(survey) - old_n))
-  follow_up2(survey)
+  expect_warning(expect_warning(expect_warning(fu3 <- follow_up2(survey, iterations = 10))))
 
   # old with batches and confounding effect vs new
   survey$batch <- ifelse(survey$Clap %in% "Right", "old", NA)
   bn <- batch_names(create_subset(sum(survey$Clap %in% "Right", na.rm = TRUE), 20))
-  survey$batch[survey$batch == "old"] <- bn
-  expect_error(follow_up2(survey))
+  survey$batch[survey$batch %in% "old"] <- bn
+  expect_warning(expect_warning(expect_warning(fu4 <- follow_up2(survey, iterations = 10))))
 
 })

### 2023-01-27 tests/testthat/test-follow_up.R
Explain the internals, fix the test

@@ -10,7 +10,13 @@ test_that("follow_up2 works", {
   old_n <- 118
   # old vs new
   survey$batch <- c(rep("old", old_n), rep(NA, nrow(survey) - old_n))
-  expect_snapshot(fu1 <- follow_up2(survey))
+  expect_warning(
+    expect_warning(
+      expect_warning(
+        fu1 <- follow_up2(survey),
+        "with the data."),
+      "with the new samples"),
+    "some problems with the new data.")
 
   # old vs new with counfounding effects
   survey$batch <- ifelse(survey$Clap %in% "Right", "old", NA)

### 2023-02-06 R/follow_up.R
Improve links to functions

@@ -8,7 +8,7 @@
 #' it matches the name of a column in original it will be used to find previous
 #' batches.
 #' @inheritParams design
-#' @seealso [`follow_up2()`]
+#' @seealso [follow_up2()]
 #' @return  A `data.frame` with the common columns of data, a new column
 #' `old_new`, and a batch column filled with the new batches needed.
 #' @export
@@ -62,7 +62,7 @@ follow_up <- function(original, follow_up, size_subset, omit = NULL,
 #' @param batch_column The name of the column of `all_data` with the batches used.
 #' If NA it is interpreted as a new data, if not empty it is considered a batch.
 #' @inheritDotParams design -pheno
-#' @seealso `follow_up()`
+#' @seealso [follow_up()]
 #' @return A `data.frame` with the `batch_column` filled with the new batches needed.
 #' @export
 #'
@@ -153,10 +153,10 @@ follow_up2 <- function(all_data, batch_column = "batch", ...) {
 #' @param old_data,new_data A data.frame with the old and new data respectively.
 #' @inheritParams follow_up
 #' @inheritParams follow_up2
-#' @seealso `check_data()`
+#' @seealso [check_data()]
 #' @export
 #' @return Called by its side effects of warnings, but returns a logical value
-#' if there are some issues (FALSE) or not (TRUE)
+#' if there are some issues (`FALSE`) or not (`TRUE`)
 #' @examples
 #' data(survey, package = "MASS")
 #' survey1 <- survey[1:118, ]

### 2023-02-06 R/spatial.R

@@ -1,11 +1,11 @@
 #' Distribute the sample on the plate
 #'
-#' This function assumes that to process the batch the samples are distributes in
-#' a plate. Sometimes you know in advance the
+#' This function assumes that to process the batch the samples are distributed in
+#' a plate with a grid scheme.
 #' @inheritParams design
 #' @param index A list with the samples on each subgroup, as provided from
-#' `design()` or `replicates()`.
-#' @param remove_positions Character, name of positions.
+#' [design()] or [replicates()].
+#' @param remove_positions Character, name of positions to be avoided in the grid.
 #' @param rows Character, name of the rows to be used.
 #' @param columns Character, name of the rows to be used.
 #' @return The indices of which samples go with which batch.

### 2023-02-06 R/indexing.R
Add comparing index

@@ -103,3 +103,54 @@ batch_names <- function(i) {
   names <- rep(names(i), lengths(i))
   names[order(unlist(i, use.names = FALSE))]
 }
+
+#' Compares two indexes
+#'
+#' Compare the distribution of samples with two different batches.
+#' @param index1,index2 A list with the index for each sample, the name of the
+#' column in `pheno` with the batch subset or the character .
+#' @param pheno A data.frame of the samples with the characteristics to normalize.
+#' @returns A matrix with the variables and the columns of of each batch.
+#' Negative values indicate `index1` was better.
+#' @export
+#' @seealso [check_index()]
+#' @examples
+#' index1 <- create_subset(50, 24)
+#' index2 <- batch_names(create_subset(50, 24))
+#' metadata <- expand.grid(height = seq(60, 80, 5), weight = seq(100, 300, 50),
+#'                          sex = c("Male","Female"))
+#' compare_index(metadata, index1, index2)
+compare_index <- function(pheno, index1, index2) {
+  if (is.character(index1) && !index1 %in% colnames(pheno)) {
+    index1 <- use_index(index1)
+  } else if (is.character(index1) && index1 %in% colnames(pheno)) {
+    index1 <- use_index(pheno[[index1]])
+  }
+  if (is.character(index2) && !index2 %in% colnames(pheno)) {
+    index2 <- use_index(index2)
+  } else if (is.character(index2) && index2 %in% colnames(pheno)) {
+    index2 <- use_index(pheno[[index2]])
+  }
+  if (sum(lengths(index1)) != nrow(pheno)) {
+    stop("Indices do not match the number of samples in pheno.")
+  }
+  if (sum(lengths(index1)) != sum(lengths(index2))) {
+    stop("Indices don't seem from the same data, their numbers are not equivalent.")
+  }
+  if (length(index1) != length(index2)) {
+    stop("Different number of batches in the indices.")
+  }
+
+  batches <- length(index1)
+  num <- is_num(pheno)
+  eval_n <- evaluations(num)
+
+  original_pheno <- .evaluate_orig(pheno, num)
+  original_pheno["na", ] <- original_pheno["na", ]/batches
+
+  ci1 <- .check_index(index1, pheno, num, eval_n, original_pheno)
+  ci2 <- .check_index(index2, pheno, num, eval_n, original_pheno)
+
+  ci1 - ci2
+
+}

### 2023-03-22 R/QC.R
Add more robustness in argument checking

@@ -67,11 +67,11 @@ extreme_cases <- function(pheno, size, omit = NULL, iterations = 500) {
 #' QC_samplesBatch <- qcSubset(index, 10, TRUE)
 qcSubset <- function(index, size, each = FALSE) {
 
-  if (!is.logical(each)) {
+  if (!is_logical(each)) {
     stop("each should be either TRUE or FALSE")
   }
 
-  if (!is.numeric(size)) {
+  if (!is_numeric(size)) {
     stop("size should be a numeric value")
   }
 

### 2023-03-22 R/designer.R

@@ -20,9 +20,9 @@
 #' index
 design <- function(pheno, size_subset, omit = NULL, iterations = 500,
                    name = "SubSet") {
-  stopifnot(is.numeric(size_subset) && all(is.finite(size_subset)) && !any(is.na(size_subset)))
+  stopifnot(is_numeric(size_subset))
   stopifnot(length(dim(pheno)) == 2)
-  stopifnot(is.numeric(iterations) && is.finite(iterations) && !any(is.na(size_subset)))
+  stopifnot(is_numeric(iterations) && is_numeric(size_subset))
   stopifnot(is.character(name))
   .design(pheno, size_subset, omit, iterations, name, check = TRUE)
 }
@@ -111,9 +111,9 @@ design <- function(pheno, size_subset, omit = NULL, iterations = 500,
 #' head(index)
 replicates <- function(pheno, size_subset, controls, omit = NULL,
                        iterations = 500){
-  stopifnot(is.numeric(size_subset) && length(size_subset) == 1 && is.finite(size_subset) && !is.na(size_subset))
+  stopifnot(is_numeric(size_subset) && length(size_subset) == 1)
   stopifnot(length(dim(pheno)) == 2)
-  stopifnot(is.numeric(iterations) && is.finite(iterations))
+  stopifnot(is_numeric(iterations))
 
   size_data <- nrow(pheno)
   if (size_subset >= size_data) {

### 2023-03-22 R/optimum.R

@@ -51,9 +51,6 @@ sizes_batches <- function(size_data, size_subset, batches) {
   check_number(size_data, "size_data")
   check_number(size_subset, "size_subset")
   check_number(batches, "batches")
-  if (batches == 1) {
-    stop("There should be more than one batch.", call. = FALSE)
-  }
   if  (size_subset*batches < size_data) {
     stop("batches or size_subset is too small to fit all the samples.",
          call. = FALSE)
@@ -114,8 +111,3 @@ internal_batches <- function(size_data, size_subset, batches) {
   sort(out, decreasing = TRUE)
 }
 
-check_number <- function(x, name) {
-  if (length(x) != 1 || !is.numeric(x) || x <= 1) {
-    stop(name, " must be a single number bigger than 1.", call. = FALSE)
-  }
-}

### 2023-03-22 R/spatial.R

@@ -20,7 +20,7 @@ spatial <- function(index, pheno, omit = NULL, remove_positions = NULL, rows = L
          columns = 1:10, iterations = 500) {
 
   stopifnot(length(dim(pheno)) == 2)
-  stopifnot(is.numeric(iterations) && is.finite(iterations))
+  stopifnot(is_numeric(iterations))
 
   nrow <- length(rows)
   ncol <- length(columns)

### 2023-03-22 R/utils.R

@@ -87,6 +87,15 @@ is_logical <- function(x){
   isTRUE(x) || isFALSE(x)
 }
 
+is_numeric <- function(x) {
+  all(x > 0 & as.integer(x) == x & is.finite(x) & !is.na(x))
+}
+
+check_number <- function(x, name) {
+  if (length(x) != 1 || !is_numeric(x) || x <= 1) {
+    stop(sQuote(name, FALSE), " must be a single number bigger than 1.", call. = FALSE)
+  }
+}
 
 mean_difference <- function(differences, subset_ind, eval_n) {
   # Calculate the score for each subset by variable

### 2023-03-22 tests/testthat/test-optimum.R

@@ -15,7 +15,7 @@ test_that("sizes_batches", {
   expect_true(all(sizes_batches(size_data, size_batch, batches) <= size_batch))
   batches <- 1
   expect_error(sizes_batches(size_data, size_batch, batches),
-               "batches must be a single number bigger than 1.")
+               "'batches' must be a single number bigger than 1.")
 })
 
 test_that("internal_batches", {

### 2023-03-22 README.Rmd
Add more robustness in argument checking
Fix typos/wording

@@
 [![CRAN status](https://www.r-pkg.org/badges/version/experDesign)](https://CRAN.R-project.org/package=experDesign)
-[![cran checks](https://badges.cranchecks.info/worst/experDesign.svg)](https://cran.r-project.org/web/checks/check_results_experDesign.html)
+[![CRAN checks](https://badges.cranchecks.info/worst/experDesign.svg)](https://cran.r-project.org/web/checks/check_results_experDesign.html)
 [![R-CMD-check](https://github.com/llrs/experDesign/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/llrs/experDesign/actions/workflows/R-CMD-check.yaml)
@@
 ```
-You can install the development version from [GitHub](https://github.com/) with:
+
+::: {.pkgdown-devel}
+You can install the development version of pkgdown from [GitHub](https://github.com/llrs/experDesign) with:
 
@@
 ```
+:::
 
@@
 
-The dataset has numeric, categorical values and some `NA`'s value.
+The dataset has numeric, categorical values and `NA` values.
 

### 2023-04-05 README.Rmd
Fix urls

@@
 [![R-CMD-check](https://github.com/llrs/experDesign/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/llrs/experDesign/actions/workflows/R-CMD-check.yaml)
-[![Coverage status](https://codecov.io/gh/llrs/experDesign/branch/master/graph/badge.svg)](https://codecov.io/github/llrs/experDesign?branch=master)
+[![Coverage status](https://codecov.io/gh/llrs/experDesign/branch/master/graph/badge.svg)](https://app.codecov.io/github/llrs/experDesign)
 [![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)

### 2024-05-07 R/spatial.R
Allow to remove rows and columns
Improves position handling via a helper

Close #52

@@ -22,29 +22,12 @@ spatial <- function(index, pheno, omit = NULL, remove_positions = NULL, rows = L
   stopifnot(length(dim(pheno)) == 2)
   stopifnot(is_numeric(iterations))
 
-  nrow <- length(rows)
-  ncol <- length(columns)
-
-  if (is.null(rows) || length(rows) == 0) {
-    stop("Please provide at least one row.", call. = FALSE)
-  }
-  if (is.null(columns) || length(columns) == 0) {
-    stop("Please provide at least one column.", call. = FALSE)
-  }
-  if ((nrow*ncol - length(remove_positions)) < max(lengths(index))) {
+  position <- handle_positions(rows, columns, remove_positions)
+  if (length(position) < max(lengths(index))) {
     stop("The size for the batch is smaller than the samples it must contain.",
          "\n\tPlease check the rows and columns or how you created the index.",
          call. = FALSE)
   }
-
-  positions <- position_name(rows, columns)
-  if (any(!remove_positions %in% positions$name)) {
-    stop("Unrecognized position to remove.",
-         "\n\tCheck that it is a combination of rows and columns: A1, A3...",
-         call. = FALSE)
-  }
-
-  position <- positions$name[!positions$name %in% remove_positions]
 
   opt <- Inf
 
@@ -69,7 +52,6 @@ spatial <- function(index, pheno, omit = NULL, remove_positions = NULL, rows = L
 
   eval_n <- evaluations(num)
 
-  n_positions <- length(position)
   for (j in seq_len(iterations)) {
 
     i <- create_index4index(i2, name = position)
@@ -89,5 +71,40 @@ spatial <- function(index, pheno, omit = NULL, remove_positions = NULL, rows = L
          "It allocated more sample to the previous index than possible.",
          call. = FALSE)
   }
-  val
+  # Return positions ordered by row and column
+  m <- match(position_name(rows, columns)$name, names(val))
+  val[m[!is.na(m)]]
+}
+
+handle_positions <- function(rows, columns, remove_positions) {
+
+  if (is.null(rows) || length(rows) == 0) {
+    stop("Please provide at least one row.", call. = FALSE)
+  }
+  if (is.null(columns) || length(columns) == 0) {
+    stop("Please provide at least one column.", call. = FALSE)
+  }
+
+  positions <- position_name(rows, columns)
+
+  k_position <- !positions$name %in% remove_positions
[…]

### 2024-05-07 R/utils.R

@@ -70,9 +70,9 @@ translate_index <- function(index,
 
 position_name <- function(rows, columns) {
   positions <- expand.grid(rows, columns, stringsAsFactors = FALSE)
-  positions$Var2 <- as.character(positions$Var2)
-  positions$name <- apply(positions, 1, paste0, collapse = "")
   colnames(positions)[1:2] <- c("row", "column")
+  positions <- positions[order(positions$row, positions$column), ]
+  positions$name <- paste0(positions$row, positions$column)
   positions
 }
 

### 2024-05-07 tests/testthat/test-spatial.R

@@ -39,3 +39,30 @@ test_that("spatial don't duplicate samples", {
   expect_false(any(table(batch_names(index), batch_names(index2)) > 1))
 })
 
+test_that("Removing rows works", {
+  data(survey, package = "MASS")
+  set.seed(21829)
+  nas <- c(137, 70)
+  index <- design(survey[-nas, c("Sex", "Smoke", "Age")], size_subset = 50,
+                  iterations = 25)
+  index2 <- spatial(index, survey[-nas, c("Sex", "Smoke", "Age")],
+                    rows = LETTERS[1:9], remove_positions = "B",
+                    columns = 1:12, iterations = 25)
+  expect_lte(length(index2), 9*12)
+  expect_true(all(names(index2) %in% position_name(rows = LETTERS[1:9], 1:12)$name))
+  expect_false(any(table(batch_names(index), batch_names(index2)) > 1))
+})
+
+test_that("Removing columns works", {
+  data(survey, package = "MASS")
+  set.seed(21289)
+  nas <- c(137, 70)
+  index <- design(survey[-nas, c("Sex", "Smoke", "Age")], size_subset = 50,
+                  iterations = 25)
+  index2 <- spatial(index, survey[-nas, c("Sex", "Smoke", "Age")],
+                    rows = LETTERS[1:9], remove_positions =  "A",
+                    columns = 1:12, iterations = 25)
+  expect_lte(length(index2), 9*12)
+  expect_true(all(names(index2) %in% position_name(rows = LETTERS[1:9], 1:12)$name))
+  expect_false(any(table(batch_names(index), batch_names(index2)) > 1))
+})

### 2024-05-20 README.Rmd
Export position_name

@@
 
-The goal of experDesign is to help you manage your samples before an experiment but after they are collected.
-For example, checking for common problems in the data, and help reduce or even prevent batch bias before performing an experiment, or measure once it is already done
+The goal of experDesign is to help you distribute your samples before an experiment but after they are collected. 
+For example, checking for common problems in the data, and reducing or even preventing batch bias before performing an experiment, or measuring it once the experiment is performed.
 It provides four main functions:
@@
 * `spatial()`: Randomize the samples on a spatial grid.
+
+There are other helpers.
 
@@
 If you are still designing the experiment and do not have collected any data [DeclareDesign](https://cran.r-project.org/package=DeclareDesign) might be relevant for you.
+But specially the [randomizr](https://cran.r-project.org/package=randomizr) packages  which makes common forms of random assignment and sampling.
 

### 2024-05-20 R/evaluate_category.R

@@ -79,7 +79,8 @@ evaluate_independence <- function(i, pheno) {
 #' @param pheno Data.frame with the variables of each sample, one row one sample.
 #' @param omit Character vector with the names of the columns to omit.
 #' @param na.omit Check the effects of missing values too.
-#' @return A logical value indicating if everything is alright (TRUE) or not (FALSE).
+#' @return A logical value indicating if everything is alright (`TRUE` or not (`FALSE`).
+#' @seealso [valid_followup()].
 #' @export
 #' @examples
 #' rdata <- expand.grid(sex = c("M", "F"), class = c("lower", "median", "high"))

### 2024-05-20 R/follow_up.R

@@ -16,7 +16,7 @@
 #' data(survey, package = "MASS")
 #' survey1 <- survey[1:118, ]
 #' survey2 <- survey[119:nrow(survey), ]
-#' fu <- follow_up(survey1, survey2, size_subset = 50, iterations = 10)
+#' folu <- follow_up(survey1, survey2, size_subset = 50, iterations = 10)
 follow_up <- function(original, follow_up, size_subset, omit = NULL,
                       old_new = "batch", iterations = 500) {
   stopifnot(is.character(old_new) & length(old_new) == 1)

### 2024-05-20 R/utils.R

@@ -68,7 +68,21 @@ translate_index <- function(index,
   index
 }
 
+
+#' Create position names for a grid.
+#'
+#' @param rows Names of the rows.
+#' @param columns Names of the columns.
+#'
+#' @return A data.frame with the rows and columns and the resulting name row+column.
+#' The name column is a factor for easier sorting in row, column order.
+#' @export
+#' @examples
+#' position_name(c("A", "B"), 1:2)
 position_name <- function(rows, columns) {
+  if (length(intersect(rows, columns)) >= 1) {
+    stop("rows and columns shouldn't share names.")
+  }
   positions <- expand.grid(rows, columns, stringsAsFactors = FALSE)
   colnames(positions)[1:2] <- c("row", "column")
   positions <- positions[order(positions$row, positions$column), ]

### 2024-05-20 tests/testthat/test-spatial.R

@@ -66,3 +66,12 @@ test_that("Removing columns works", {
   expect_true(all(names(index2) %in% position_name(rows = LETTERS[1:9], 1:12)$name))
   expect_false(any(table(batch_names(index), batch_names(index2)) > 1))
 })
+
+test_that("position_name works", {
+  expect_error(position_name(1:2, 1:2))
+  pn <- position_name(LETTERS[1:2], 1:2)
+  expect_s3_class(pn, "data.frame")
+  expect_equal(pn$row, rep(LETTERS[1:2], each = 2))
+  expect_equal(pn$column, rep(1:2, length.out = 4))
+  expect_equal(pn$name, as.factor(paste0(rep(LETTERS[1:2], each = 2), rep(1:2, length.out = 4))))
+})

### 2024-05-20 README.Rmd
Change branch name

@@
 [![R-CMD-check](https://github.com/llrs/experDesign/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/llrs/experDesign/actions/workflows/R-CMD-check.yaml)
-[![Coverage status](https://codecov.io/gh/llrs/experDesign/branch/master/graph/badge.svg)](https://app.codecov.io/github/llrs/experDesign)
+[![Coverage status](https://codecov.io/gh/llrs/experDesign/branch/devel/graph/badge.svg)](https://app.codecov.io/github/llrs/experDesign)
 [![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)
